---
title: "Advanced Data Operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Data Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
library(survival)
```

## Introduction

Clinical research often involves complex data manipulation tasks that go beyond basic cleaning and analysis. Real-world studies frequently require integrating data from multiple sources, tracking patient cohorts through complex eligibility criteria, and transforming data between different formats for various analytical needs.

### Common Challenges in Clinical Data Operations

In multi-center clinical studies and observational research using electronic health records (EHR), researchers commonly face:

- **Multi-source data integration**: Combining data from different systems (EMR, lab systems, registries)
- **Fuzzy matching**: Linking records when exact identifiers don't match (e.g., diagnosis descriptions)
- **Temporal alignment**: Matching events that occurred within specific time windows
- **Cohort tracking**: Documenting how many patients were excluded at each study stage
- **Format conversion**: Transforming between long and wide formats for different analyses

### Real-World Scenarios

These challenges arise in scenarios such as:

1. **Retrospective cohort studies**: Identifying eligible patients from EHR data with complex inclusion/exclusion criteria
2. **Multi-site trials**: Merging data from different hospitals using different coding systems
3. **Longitudinal analyses**: Tracking patients across multiple visits with varying assessment schedules
4. **Registry linkage**: Combining clinical trial data with national death registries or insurance claims

The `clinpubr` package provides advanced tools for these sophisticated data manipulation workflows.

## Cohort Screening with Exclusion Tracking

### Building Study Cohorts from EHR Data

In observational research, identifying the right patient cohort is critical. Unlike randomized trials where patients are prospectively enrolled, retrospective studies must:

1. Start with a broad population
2. Apply inclusion criteria (e.g., specific diagnoses, age ranges)
3. Apply exclusion criteria (e.g., comorbidities, missing data)
4. **Track how many patients are lost at each step** (CONSORT-style documentation)

Clinical studies must document how the final analysis cohort was derived. This is required by:
- **CONSORT guidelines**: For randomized trials
- **STROBE guidelines**: For observational studies
- **Regulatory submissions**: FDA, EMA require clear documentation

The `exclusion_count()` function creates a standardized flow diagram automatically.

```{r cohort-screening}
# Create sample multi-table data representing a typical EHR structure
# In real studies, these might come from different database tables

set.seed(123)

# Patient demographics table
patients <- data.frame(
  patient_id = 1:100,
  age = round(rnorm(100, 60, 15)),
  gender = sample(c("M", "F"), 100, replace = TRUE),
  enrollment_date = Sys.Date() - sample(1:365, 100, replace = TRUE),
  has_baseline_data = sample(c(TRUE, FALSE), 100, prob = c(0.9, 0.1), replace = TRUE),
  meets_inclusion = sample(c(TRUE, FALSE), 100, prob = c(0.8, 0.2), replace = TRUE),
  complete_followup = sample(c(TRUE, FALSE), 100, prob = c(0.85, 0.15), replace = TRUE),
  stringsAsFactors = FALSE
)

# View initial cohort
cat("Initial cohort:", nrow(patients), "patients\n")
knitr::kable(head(patients[, c("patient_id", "age", "gender", "has_baseline_data")]), 
             caption = "Sample of Initial Cohort Data")
```

### Multi-stage Cohort Screening

Apply multiple criteria sequentially and track exclusions at each stage:

```{r multi-stage-screening}
# Track exclusions using exclusion_count
# This creates a CONSORT-style flow diagram automatically

exclusion_summary <- exclusion_count(
  patients,
  age < 50,                          # Age criterion: adults >= 50
  gender != "M",                     # Gender criterion: male only
  !has_baseline_data,                # Data completeness
  !meets_inclusion,                  # Study-specific inclusion
  !complete_followup,                # Follow-up completeness
  .criteria_names = c(
    "Age < 50 years",
    "Not male",
    "Missing baseline data",
    "Does not meet inclusion criteria",
    "Incomplete follow-up"
  )
)

knitr::kable(exclusion_summary, caption = "CONSORT-Style Cohort Flow")
```

### Interpreting the Flow Diagram

The `exclusion_count()` output shows:

- **Step**: Each screening criterion in order
- **N**: Number of patients remaining after each criterion
- **Excluded**: Number of patients excluded at each step
- **Criteria**: Description of the exclusion criterion

```{r screening-summary}
# Calculate retention statistics
initial_n <- exclusion_summary$N[1]
final_n <- exclusion_summary$N[nrow(exclusion_summary)]
retention_rate <- round(100 * final_n / initial_n, 1)

cat("\n=== Cohort Screening Summary ===\n")
cat("Initial cohort:", initial_n, "\n")
cat("Final analysis cohort:", final_n, "\n")
cat("Total excluded:", initial_n - final_n, "\n")
cat("Retention rate:", retention_rate, "%\n")
```

### Customizing Exclusion Criteria

You can customize the exclusion criteria to match your specific study needs:

```{r exclusion-custom}
# Example: Age-restricted study with different criteria
exclusion_flow <- exclusion_count(
  patients,
  age < 18 | age > 80,               # Age range criterion
  !has_baseline_data,                # Data completeness
  !meets_inclusion,                  # Study-specific inclusion
  .criteria_names = c(
    "Age outside 18-80 range",
    "Missing baseline data",
    "Does not meet inclusion criteria"
  )
)

knitr::kable(exclusion_flow, caption = "Custom Exclusion Flow")
```

## Data Merging Strategies

### Merge by Substring Matching

In real-world data, exact matching is often impossible. For example:
- **Diagnosis coding**: Different hospitals may use different terminology ("Type 2 Diabetes" vs "Diabetes Mellitus Type 2")
- **Medication names**: Brand names vs generic names, or different abbreviations
- **Procedure descriptions**: Varying levels of detail in procedure notes

The `merge_by_substring()` function enables fuzzy matching based on text similarity.

```{r merge-substring}
# Example: Mapping free-text diagnoses to standardized ICD codes
# This is common when working with EHR data from multiple sources

# Standard reference table (e.g., from a coding system)
medical_terms <- data.frame(
  term = c("Diabetes Mellitus Type 2", "Hypertension", "Coronary Artery Disease",
           "Chronic Kidney Disease", "Heart Failure", "Atrial Fibrillation"),
  icd_code = c("E11", "I10", "I25", "N18", "I50", "I48"),
  category = c("Endocrine", "Cardiovascular", "Cardiovascular", 
               "Renal", "Cardiovascular", "Cardiovascular"),
  stringsAsFactors = FALSE
)

# Patient diagnoses from EHR (free text, potentially messy)
patient_diagnoses <- data.frame(
  patient_id = 1:10,
  diagnosis_text = c(
    "Type 2 Diabetes Mellitus",           # Exact match conceptually, different word order
    "Essential Hypertension",              # Partial match
    "CAD - Coronary Artery Disease",       # Abbreviation + full text
    "CKD Stage 3",                         # Abbreviation
    "Congestive Heart Failure",            # Synonym
    "Diabetes Type 2",                     # Different word order
    "Hypertensive disease",                # Different terminology
    "Coronary disease",                    # Partial match
    "Kidney disease chronic",              # Different word order
    "Heart failure chronic"                # Different word order
  ),
  stringsAsFactors = FALSE
)

# Show the challenge: free-text vs standardized codes
cat("Sample diagnosis mapping challenge:\n")
knitr::kable(head(patient_diagnoses, 5), caption = "Free-Text Diagnoses (Before Mapping)")

# Merge by substring matching
merged_substring <- merge_by_substring(
  data = patient_diagnoses,
  key_df = medical_terms,
  search_col = "diagnosis_text",
  key_col = "term",
  value_cols = c("icd_code", "category")
)

cat("\nAfter substring matching:\n")
knitr::kable(head(merged_substring), caption = "Diagnoses with Mapped ICD Codes")

# Summary of mapping success
mapping_summary <- data.frame(
  Metric = c("Total records", "Successfully mapped", "Mapping rate (%)"),
  Value = c(
    nrow(patient_diagnoses),
    sum(!is.na(merged_substring$icd_code)),
    round(100 * sum(!is.na(merged_substring$icd_code)) / nrow(patient_diagnoses), 1)
  )
)
knitr::kable(mapping_summary, caption = "Mapping Success Rate")
```

### Merge by Range (Time-Based Matching)

In longitudinal studies, we often need to match events that occurred within specific time windows. Examples include:
- **Lab values during hospitalization**: Match lab results to specific hospital stays
- **Medication dispensing**: Link prescriptions to clinic visits
- **Adverse events**: Associate side effects with treatment periods

```{r merge-range}
# Example: Match lab results to hospital visits
# This is common when analyzing inpatient data

# Patient visit records (admission and discharge dates)
patient_visits <- data.frame(
  patient_id = rep(1:3, each = 2),
  visit_id = 1:6,
  visit_start = as.Date(c("2023-01-01", "2023-06-01",
                          "2023-02-01", "2023-07-01",
                          "2023-03-01", "2023-08-01")),
  visit_end = as.Date(c("2023-01-10", "2023-06-10",
                        "2023-02-10", "2023-07-10",
                        "2023-03-10", "2023-08-10")),
  visit_type = c("Admission", "Follow-up", "Admission", "Follow-up", 
                 "Admission", "Follow-up"),
  stringsAsFactors = FALSE
)

# Lab results with collection dates
lab_results <- data.frame(
  lab_id = 1:6,
  patient_id = c(1, 1, 2, 2, 3, 3),
  test_date = as.Date(c("2023-01-05", "2023-06-05",
                        "2023-02-03", "2023-07-08",
                        "2023-03-15", "2023-08-05")),
  test_name = c("Glucose", "HbA1c", "Glucose", "Creatinine", "Glucose", "Lipid panel"),
  result = round(rnorm(6, 100, 15), 1),
  stringsAsFactors = FALSE
)

# Show the data before merging
cat("Patient visits:\n")
knitr::kable(patient_visits, caption = "Hospital Visits")

cat("\nLab results:\n")
knitr::kable(lab_results, caption = "Laboratory Results")

# Merge by range: find lab results within each visit window
merged_range <- merge_by_range(
  x = patient_visits,
  y = lab_results,
  by = "patient_id",
  x_start = "visit_start",
  x_end = "visit_end",
  y_val = "test_date"
)

cat("\nMatched lab results to visits:\n")
knitr::kable(merged_range, caption = "Labs Matched to Visit Windows")
```

## Data Transformation

### Long to Wide Format Conversion

Clinical data is often stored in long format (one row per measurement), but many analyses require wide format (one row per patient with multiple columns for measurements).

**When to use wide format:**
- **Repeated measures ANOVA**: Each time point as a separate column
- **Change score calculations**: Baseline vs follow-up comparisons
- **Pattern identification**: Visualizing trajectories across visits

```{r to-wide}
# Create long-format lab data (typical EHR structure)
long_labs <- data.frame(
  patient_id = rep(1:5, each = 3),
  visit = rep(1:3, 5),
  test = rep(c("glucose", "creatinine", "cholesterol"), 5),
  value = round(rnorm(15, 100, 20), 1),
  stringsAsFactors = FALSE
)

cat("Long format (typical EHR structure):\n")
knitr::kable(head(long_labs, 8), caption = "Laboratory Data - Long Format")

# Convert to wide format for analysis
wide_labs <- to_wide(
  df = long_labs,
  keys = c("patient_id", "visit"),
  item_col = "test",
  value_col = "value"
)

cat("\nWide format (for analysis):\n")
knitr::kable(head(wide_labs), caption = "Laboratory Data - Wide Format")
```

### Working with Multi-choice Data

Survey data and clinical assessments often use multi-choice questions (e.g., "Select all symptoms that apply"). These need special handling:

```{r multichoice-advanced}
# Create complex multi-choice survey data
# Common in patient-reported outcomes and symptom assessments
set.seed(456)
survey_data <- data.frame(
  respondent_id = 1:20,
  symptoms = sapply(1:20, function(x) {
    paste(sample(c("fever", "cough", "headache", "fatigue", "nausea"), 
                 sample(1:3, 1)), collapse = ",")
  }),
  comorbidities = sapply(1:20, function(x) {
    paste(sample(c("diabetes", "hypertension", "asthma", "none"), 
                 sample(1:2, 1)), collapse = ",")
  }),
  stringsAsFactors = FALSE
)

cat("Original multi-choice format:\n")
knitr::kable(head(survey_data), caption = "Multi-Choice Survey Data")

# Split multi-choice columns into binary indicators
# This creates one column per option (TRUE/FALSE)
symptoms_split <- split_multichoice(
  survey_data,
  quest_cols = c("symptoms", "comorbidities"),
  split = ",",
  remove_space = FALSE
)

cat("\nAfter splitting (binary indicators):\n")
knitr::kable(head(symptoms_split), caption = "Split Multi-Choice Data")

# Combine related symptoms into composite variables
# Useful for creating clinically meaningful groupings
combined <- combine_multichoice(
  symptoms_split,
  quest_cols = list(
    respiratory = c("symptoms_cough", "symptoms_fatigue"),
    systemic = c("symptoms_fever", "symptoms_headache")
  ),
  sep = ","
)

cat("\nCombined symptom groups:\n")
knitr::kable(head(data.frame(
  respondent_id = symptoms_split$respondent_id,
  respiratory_symptoms = combined$respiratory,
  systemic_symptoms = combined$systemic
)), caption = "Combined Symptom Groups")
```

## Utility Functions

### String Manipulation for Data Standardization

```{r string-utils}
# Find common prefix in file names
# Useful when processing batches of files with similar naming
file_names <- c("patient_001_lab.csv", "patient_001_visit.csv", "patient_001_demo.csv")
common_prefix(file_names)

# Standardize test names across different lab systems
# Different labs may use different naming conventions
test_names <- c("Glucose_Fasting", "Glucose_Random", "Cholesterol_Total", "LDL_Calculated")
standardized_names <- str_match_replace(
  x = test_names,
  to_match = c("Glucose", "Cholesterol", "LDL"),
  to_replace = c("GLU", "CHOL", "LDL_CHOL")
)

knitr::kable(data.frame(
  Original = test_names,
  Standardized = standardized_names
), caption = "Test Name Standardization")
```

### List Operations for Complex Data Structures

```{r list-ops}
# Add lists element-wise
# Useful for aggregating counts across multiple categories
list1 <- list(diabetes = 10, hypertension = 20, asthma = 5)
list2 <- list(diabetes = 15, hypertension = 25, asthma = 8)
combined_counts <- add_lists(list1, list2)

cat("Combined condition counts:\n")
print(combined_counts)

# Merge ordered vectors preserving order
# Useful for creating master lists of medications or diagnoses
site1_conditions <- c("Diabetes", "Hypertension", "Heart Failure", "CKD")
site2_conditions <- c("Hypertension", "COPD", "Diabetes", "Cancer")
site3_conditions <- c("CKD", "Diabetes", "Stroke", "Hypertension")

master_list <- merge_ordered_vectors(list(site1_conditions, site2_conditions, site3_conditions))
cat("\nMaster condition list (ordered by first appearance):\n")
print(master_list)
```

### Data Frame Operations

```{r df-ops}
# Replace elements based on conditions
# Common for recoding categorical variables
sample_data <- data.frame(
  group = c("A", "B", "A", "C", "B"),
  value = c(1, 2, 3, 4, 5),
  stringsAsFactors = FALSE
)

# Recode groups to descriptive labels
recoded <- replace_elements(
  x = sample_data$group,
  from = c("A", "B", "C"),
  to = c("Control", "Treatment", "Placebo")
)

cat("Original vs recoded:\n")
knitr::kable(data.frame(
  Original = sample_data$group,
  Recoded = recoded
), caption = "Group Recoding")

# Fill NA values with last valid value
# Useful for carrying forward last observation in longitudinal data
time_series <- c(120, NA, NA, 125, NA, 130, NA, NA)
filled_series <- fill_with_last(time_series)

cat("\nMissing data imputation (LOCF):\n")
knitr::kable(data.frame(
  Original = time_series,
  Filled = filled_series
), caption = "Last Observation Carried Forward")
```

## Complete Data Processing Workflow

Here's a comprehensive workflow combining multiple advanced operations for a realistic clinical research scenario:

```{r complete-workflow}
# Scenario: Building an analysis cohort for a diabetes study
# Data sources: Demographics, visits, labs from different EHR tables

cat("=== Clinical Data Processing Workflow ===\n\n")

# Step 1: Create sample multi-source clinical data
cat("Step 1: Creating sample multi-source clinical data...\n")

# Demographics table
demographics <- data.frame(
  patient_id = 1:50,
  age = sample(25:75, 50, replace = TRUE),
  gender = sample(c("M", "F"), 50, replace = TRUE),
  enrollment_date = Sys.Date() - sample(1:365, 50, replace = TRUE),
  has_baseline_data = sample(c(TRUE, FALSE), 50, prob = c(0.9, 0.1), replace = TRUE),
  meets_inclusion = sample(c(TRUE, FALSE), 50, prob = c(0.8, 0.2), replace = TRUE),
  stringsAsFactors = FALSE
)

# Visits table (some patients have multiple visits)
visits <- data.frame(
  visit_id = 1:150,
  patient_id = sample(1:50, 150, replace = TRUE),
  visit_date = Sys.Date() - sample(1:200, 150, replace = TRUE),
  bp_systolic = round(rnorm(150, 130, 15)),
  bp_diastolic = round(rnorm(150, 85, 10)),
  stringsAsFactors = FALSE
)

# Lab results table
labs <- data.frame(
  lab_id = 1:200,
  patient_id = sample(1:50, 200, replace = TRUE),
  collection_date = Sys.Date() - sample(1:180, 200, replace = TRUE),
  test_type = sample(c("glucose", "hba1c", "creatinine", "ldl"), 200, replace = TRUE),
  result = round(rnorm(200, 100, 25), 2),
  stringsAsFactors = FALSE
)

cat("Data sources created:\n")
cat("- Demographics:", nrow(demographics), "patients\n")
cat("- Visits:", nrow(visits), "records\n")
cat("- Labs:", nrow(labs), "results\n")

# Step 2: Screen for eligible patients with CONSORT-style tracking
cat("\nStep 2: Screening for eligible patients with exclusion tracking...\n")

final_cohort <- exclusion_count(
  demographics,
  age < 40,                          # Age criterion
  !has_baseline_data,                # Data completeness
  !meets_inclusion,                  # Study-specific inclusion
  .criteria_names = c(
    "Age < 40 years",
    "Missing baseline data",
    "Does not meet inclusion criteria"
  )
)

knitr::kable(final_cohort, caption = "Cohort Screening Flow")

# Step 3: Merge visits with closest lab results
cat("\nStep 3: Merging data sources...\n")

# For each visit, find labs collected within a window
merged_data <- merge_by_range(
  x = visits,
  y = labs,
  by = "patient_id",
  x_start = "visit_date",
  x_end = "visit_date",
  y_val = "collection_date",
  suffixes = c("_visit", "_lab")
)

cat("Merged records (visits + labs):", nrow(merged_data), "\n")

# Step 4: Transform to wide format for analysis
cat("\nStep 4: Transforming data...\n")

# Summarize labs by patient and test type (average results)
lab_summary <- labs %>%
  group_by(patient_id, test_type) %>%
  summarise(avg_result = mean(result, na.rm = TRUE), .groups = "drop")

# Convert to wide format: one row per patient, columns for each test
lab_wide <- to_wide(
  df = lab_summary,
  keys = "patient_id",
  item_col = "test_type",
  value_col = "avg_result"
)

cat("Wide format dimensions:", nrow(lab_wide), "rows,", ncol(lab_wide), "columns\n")
knitr::kable(head(lab_wide, 5), caption = "Lab Data - Wide Format Summary")

# Summary statistics
cat("\n=== Workflow Summary ===\n")
cat("Initial patients:", nrow(demographics), "\n")
cat("Final analysis cohort:", final_cohort$N[nrow(final_cohort)], "\n")
cat("Retention rate:", round(100 * final_cohort$N[nrow(final_cohort)] / nrow(demographics), 1), "%\n")

cat("\nWorkflow complete!\n")
```

## Summary

The `clinpubr` package provides powerful tools for advanced data operations essential in clinical research:

### Key Functions and Their Applications

| Function | Clinical Application |
|----------|---------------------|
| `exclusion_count()` | CONSORT-style cohort flow documentation |
| `merge_by_substring()` | Fuzzy matching for diagnosis/procedure coding |
| `merge_by_range()` | Time-window matching for longitudinal data |
| `to_wide()` | Format conversion for repeated measures analysis |
| `split_multichoice()` / `combine_multichoice()` | Survey and symptom data processing |
| **Utility functions** | Data standardization and cleaning |

### Best Practices

1. **Document your cohort flow**: Always use `exclusion_count()` to track how your final cohort was derived
2. **Validate fuzzy matching**: When using `merge_by_substring()`, manually review a sample of matches
3. **Check time windows**: Ensure your range matching captures the intended temporal relationships
4. **Handle missing data explicitly**: Don't silently drop missing values - document them
5. **Standardize early**: Apply string standardization and recoding at the beginning of your workflow

### Common Pitfalls to Avoid

- **Over-matching with substrings**: Can create false matches; validate with manual review
- **Overlapping time windows**: Be clear about whether boundaries are inclusive or exclusive
- **Silent data loss**: Always check how many records are lost at each merge step
- **Format confusion**: Document whether your data is in long or wide format at each stage

These functions enable sophisticated data manipulation workflows essential for complex clinical research projects, from retrospective EHR studies to multi-site clinical trials.

## Further Reading

- `vignette("data-cleaning")`: Data cleaning and preparation
- `vignette("baseline-table")`: Creating baseline characteristics tables
- `vignette("regression-analysis")`: Regression analysis and visualization
- `vignette("model-evaluation")`: Model comparison and evaluation
