---
title: "Advanced Data Operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Data Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5,
    warning = FALSE,
    message = FALSE,
    error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
library(survival)
```

## Introduction

Clinical research often involves complex data manipulation tasks that go beyond basic cleaning and analysis. Real-world studies frequently require integrating data from multiple sources, tracking patient cohorts through complex eligibility criteria, and transforming data between different formats for various analytical needs.

### Common Challenges in Clinical Data Operations

In multi-center clinical studies and observational research using electronic health records (EHR), researchers commonly face:

- **Multi-source data integration**: Combining data from different systems (EMR, lab systems, registries)
- **Fuzzy matching**: Linking records when exact identifiers don't match (e.g., diagnosis descriptions)
- **Temporal alignment**: Matching events that occurred within specific time windows
- **Cohort tracking**: Documenting how many patients were excluded at each study stage
- **Format conversion**: Transforming between long and wide formats for different analyses

### Real-World Scenarios

These challenges arise in scenarios such as:

1. **Retrospective cohort studies**: Identifying eligible patients from EHR data with complex inclusion/exclusion criteria
2. **Multi-site trials**: Merging data from different hospitals using different coding systems
3. **Longitudinal analyses**: Tracking patients across multiple visits with varying assessment schedules
4. **Registry linkage**: Combining clinical trial data with national death registries or insurance claims

The `clinpubr` package provides advanced tools for these sophisticated data manipulation workflows.

## Cohort Screening with Exclusion Tracking

### Building Study Cohorts from EHR Data

In observational research, identifying the right patient cohort is critical. Unlike randomized trials where patients are prospectively enrolled, retrospective studies must:

1. Start with a broad population
2. Apply inclusion criteria (e.g., specific diagnoses, age ranges)
3. Apply exclusion criteria (e.g., comorbidities, missing data)
4. **Track how many patients are lost at each step** (CONSORT-style documentation)

Clinical studies must document how the final analysis cohort was derived. This is required by:
- **CONSORT guidelines**: For randomized trials
- **STROBE guidelines**: For observational studies
- **Regulatory submissions**: FDA, EMA require clear documentation

The `exclusion_count()` function creates a standardized flow diagram automatically.

First, create sample multi-table data representing a typical EHR structure. In real studies, these might come from different database tables.

```{r cohort-screening-setup}
set.seed(123)

# Patient demographics table
patients <- data.frame(
    patient_id = 1:100,
    age = round(rnorm(100, 60, 15)),
    gender = sample(c("M", "F"), 100, replace = TRUE),
    enrollment_date = Sys.Date() - sample(1:365, 100, replace = TRUE),
    has_baseline_data = sample(c(TRUE, FALSE), 100, prob = c(0.9, 0.1), replace = TRUE),
    meets_inclusion = sample(c(TRUE, FALSE), 100, prob = c(0.8, 0.2), replace = TRUE),
    complete_followup = sample(c(TRUE, FALSE), 100, prob = c(0.85, 0.15), replace = TRUE),
    stringsAsFactors = FALSE
)
```

View the initial cohort:

```{r cohort-screening-view}
cat("Initial cohort:", nrow(patients), "patients\n")
knitr::kable(head(patients),
    caption = "Sample of Initial Cohort Data"
)
```

### Multi-stage Cohort Screening

Apply multiple criteria sequentially and track exclusions at each stage. The `exclusion_count()` function creates a CONSORT-style flow diagram automatically.

```{r multi-stage-screening}
exclusion_summary <- exclusion_count(
    patients,
    age < 50, # Age criterion: adults >= 50
    gender != "M", # Gender criterion: male only
    !has_baseline_data, # Data completeness
    !meets_inclusion, # Study-specific inclusion
    !complete_followup # Follow-up completeness
)

knitr::kable(exclusion_summary, caption = "CONSORT-Style Cohort Flow")
```

### Interpreting the Flow Diagram

The `exclusion_count()` output shows:

- **Step**: Each screening criterion in order
- **N**: Number of patients remaining after each criterion
- **Excluded**: Number of patients excluded at each step
- **Criteria**: Description of the exclusion criterion

Calculate retention statistics:

```{r screening-summary}
initial_n <- exclusion_summary$N[1]
final_n <- exclusion_summary$N[nrow(exclusion_summary)]
retention_rate <- round(100 * final_n / initial_n, 1)

cat("\n=== Cohort Screening Summary ===\n")
cat("Initial cohort:", initial_n, "\n")
cat("Final analysis cohort:", final_n, "\n")
cat("Total excluded:", initial_n - final_n, "\n")
cat("Retention rate:", retention_rate, "%\n")
```

### Customizing Exclusion Criteria

You can customize the exclusion criteria to match your specific study needs. Here is an example for an age-restricted study with different criteria:

```{r exclusion-custom}
exclusion_flow <- exclusion_count(
    patients,
    age < 18 | age > 80, # Age range criterion
    !has_baseline_data, # Data completeness
    !meets_inclusion, # Study-specific inclusion
    .criteria_names = c(
        "Age outside 18-80 range",
        "Missing baseline data",
        "Does not meet inclusion criteria"
    )
)

knitr::kable(exclusion_flow, caption = "Custom Exclusion Flow")
```

## Data Merging Strategies

### Merge by Substring Matching

In real-world data, exact matching is often impossible. For example:
- **Diagnosis coding**: Different hospitals may use different terminology ("Type 2 Diabetes" vs "Diabetes Mellitus Type 2")
- **Medication names**: Brand names vs generic names, or different abbreviations
- **Procedure descriptions**: Varying levels of detail in procedure notes

The `merge_by_substring()` function enables matching based on substring containment. It checks if the `search_col` string contains the `key_col` string as a substring.

```{r merge-substring}
# Example: Mapping free-text diagnoses to standardized ICD codes
# This is common when working with EHR data from multiple sources

# Standard reference table (e.g., from a coding system)
medical_terms <- data.frame(
  term = c("Type 2 Diabetes", "Hypertension", "Coronary Artery Disease",
           "Chronic Kidney Disease", "Heart Failure", "Atrial Fibrillation"),
  icd_code = c("E11", "I10", "I25", "N18", "I50", "I48"),
  category = c("Endocrine", "Cardiovascular", "Cardiovascular", 
               "Renal", "Cardiovascular", "Cardiovascular"),
  stringsAsFactors = FALSE
)

# Patient diagnoses from EHR (free text, potentially messy)
patient_diagnoses <- data.frame(
  patient_id = 1:10,
  diagnosis_text = c(
    "Severe Type 2 Diabetes",              # "Type 2 Diabetes" is substring
    "Type 2 Diabetes Mellitus",            # "Type 2 Diabetes" is substring
    "Type 2 Diabetes with Complications",  # "Type 2 Diabetes" is substring
    "Essential Hypertension",              # "Hypertension" is substring
    "Hypertensive disease",                # "Hypertension" is substring
    "CAD - Coronary Artery Disease",       # "Coronary Artery Disease" is substring
    "Coronary disease",                    # Partial match - may not work
    "CKD Stage 3",                         # "Chronic Kidney Disease" not substring
    "Congestive Heart Failure",            # "Heart Failure" is substring
    "Heart failure chronic"                # "Heart Failure" is substring (case insensitive)
  ),
  stringsAsFactors = FALSE
)
```

Show the challenge: free-text vs standardized codes:

```{r merge-substring-view}
cat("Sample diagnosis mapping challenge:\n")
knitr::kable(patient_diagnoses, caption = "Free-Text Diagnoses (Before Mapping)")
```

Merge by substring matching:

```{r merge-substring-apply}
merged_substring <- merge_by_substring(
  data = patient_diagnoses,
  key_df = medical_terms,
  search_col = "diagnosis_text",
  key_col = "term",
  value_cols = c("icd_code", "category")
)

cat("\nAfter substring matching:\n")
knitr::kable(head(merged_substring), caption = "Diagnoses with Mapped ICD Codes")
```

Summary of mapping success:

```{r merge-substring-summary}
mapping_summary <- data.frame(
  Metric = c("Total records", "Successfully mapped", "Mapping rate (%)"),
  Value = c(
    nrow(patient_diagnoses),
    sum(!is.na(merged_substring$icd_code)),
    round(100 * sum(!is.na(merged_substring$icd_code)) / nrow(patient_diagnoses), 1)
  )
)
knitr::kable(mapping_summary, caption = "Mapping Success Rate")
```

### Merge by Range (Time-Based Matching)

In longitudinal studies, we often need to match events that occurred within specific time windows. Examples include:
- **Lab values during hospitalization**: Match lab results to specific hospital stays
- **Medication dispensing**: Link prescriptions to clinic visits
- **Adverse events**: Associate side effects with treatment periods

First, create patient visit records (admission and discharge dates):

```{r merge-range-visits}
patient_visits <- data.frame(
    patient_id = rep(1:3, each = 2),
    visit_id = 1:6,
    visit_start = as.Date(c(
        "2023-01-01", "2023-06-01",
        "2023-02-01", "2023-07-01",
        "2023-03-01", "2023-08-01"
    )),
    visit_end = as.Date(c(
        "2023-01-10", "2023-06-10",
        "2023-02-10", "2023-07-10",
        "2023-03-10", "2023-08-10"
    )),
    visit_type = c(
        "Admission", "Follow-up", "Admission", "Follow-up",
        "Admission", "Follow-up"
    ),
    stringsAsFactors = FALSE
)

cat("Patient visits:\n")
knitr::kable(patient_visits, caption = "Hospital Visits")
```

Create lab results with collection dates:

```{r merge-range-labs}
lab_results <- data.frame(
    lab_id = 1:6,
    patient_id = c(1, 1, 2, 2, 3, 3),
    test_date = as.Date(c(
        "2023-01-05", "2023-06-05",
        "2023-02-03", "2023-07-08",
        "2023-03-15", "2023-08-05"
    )),
    test_name = c("Glucose", "HbA1c", "Glucose", "Creatinine", "Glucose", "Lipid panel"),
    result = round(rnorm(6, 100, 15), 1),
    stringsAsFactors = FALSE
)

cat("\nLab results:\n")
knitr::kable(lab_results, caption = "Laboratory Results")
```

Merge by range: find lab results within each visit window:

```{r merge-range-apply}
merged_range <- merge_by_range(
    x = patient_visits,
    y = lab_results,
    by = "patient_id",
    x_start = "visit_start",
    x_end = "visit_end",
    y_val = "test_date"
)

cat("\nMatched lab results to visits:\n")
knitr::kable(merged_range, caption = "Labs Matched to Visit Windows")
```

## Data Transformation

### Long to Wide Format Conversion

Clinical data is often stored in long format (one row per measurement), but many analyses require wide format (one row per patient with multiple columns for measurements).

**When to use wide format:**
- **Repeated measures ANOVA**: Each time point as a separate column
- **Change score calculations**: Baseline vs follow-up comparisons
- **Pattern identification**: Visualizing trajectories across visits

Create long-format lab data (typical EHR structure):

```{r to-wide-setup}
long_labs <- data.frame(
    patient_id = rep(1:5, each = 3),
    visit = c(1, 2, 1, 2, 1, 2, 2, 2, 2, 1, 2, 2, 1, 1, 1),
    test = rep(c("glucose", "creatinine", "cholesterol"), 5),
    value = round(rnorm(15, 100, 20), 1),
    stringsAsFactors = FALSE
)

cat("Long format (typical EHR structure):\n")
knitr::kable(head(long_labs, 8), caption = "Laboratory Data - Long Format")
```

Convert to wide format for analysis:

```{r to-wide-apply}
wide_labs <- to_wide(
    df = long_labs,
    keys = c("patient_id", "visit"),
    item_col = "test",
    value_col = "value"
)

cat("\nWide format (for analysis):\n")
knitr::kable(head(wide_labs), caption = "Laboratory Data - Wide Format")
```

### Working with Multi-choice Data

Survey data and clinical assessments often use multi-choice questions (e.g., "Select all symptoms that apply"). These need special handling.

Create complex multi-choice survey data (common in patient-reported outcomes and symptom assessments):

```{r multichoice-advanced-setup}
set.seed(456)
survey_data <- data.frame(
    id = 1:20,
    symptoms = sapply(1:20, function(x) {
        paste(sample(
            c("fever", "cough", "headache", "fatigue"),
            sample(1:4, 1)
        ), collapse = ",")
    }),
    comorbidities = sapply(1:20, function(x) {
        paste(sample(
            c("diabetes", "hypertension"),
            sample(1:2, 1)
        ), collapse = ",")
    }),
    stringsAsFactors = FALSE
)

cat("Original multi-choice format:\n")
knitr::kable(head(survey_data), caption = "Multi-Choice Survey Data")
```

Split multi-choice columns into binary indicators. This creates one column per option (TRUE/FALSE):

```{r multichoice-split}
symptoms_split <- split_multichoice(
    survey_data,
    quest_cols = c("symptoms", "comorbidities"),
    split = ",",
    remove_space = FALSE
)

cat("\nAfter splitting (binary indicators):\n")
knitr::kable(head(symptoms_split), caption = "Split Multi-Choice Data")
```

Combine related symptoms into composite variables. This is useful for creating clinically meaningful groupings:

```{r multichoice-combine}
combined <- combine_multichoice(
    symptoms_split,
    quest_cols = list(
        respiratory = c("symptoms_cough", "symptoms_fatigue"),
        systemic = c("symptoms_fever", "symptoms_headache")
    ),
    sep = ","
)

cat("\nCombined symptom groups:\n")
knitr::kable(head(combined), caption = "Combined Symptom Groups")
```

## Utility Functions

### String Manipulation for Data Standardization

Find common prefix in file names. This is useful when processing batches of files with similar naming:

```{r string-utils-prefix}
file_names <- c("patient_001_lab.csv", "patient_001_visit.csv", "patient_001_demo.csv")
common_prefix(file_names)
```

Standardize test names across different lab systems. Different labs may use different naming conventions:

```{r string-utils-replace}
test_names <- c("Glucose_Fasting", "Glucose_Random", "Cholesterol_Total", "LDL_Calculated")
standardized_names <- str_match_replace(
    x = test_names,
    to_match = c("Glucose", "Cholesterol", "LDL"),
    to_replace = c("GLU", "CHOL", "LDL_CHOL")
)

knitr::kable(data.frame(
    Original = test_names,
    Standardized = standardized_names
), caption = "Test Name Standardization")
```

### List Operations for Complex Data Structures

Add lists element-wise. This is useful for aggregating counts across multiple categories:

```{r list-ops-add}
list1 <- list(diabetes = 10, hypertension = 20, asthma = 5)
list2 <- list(diabetes = 15, hypertension = 25, asthma = 8)
combined_counts <- add_lists(list1, list2)

cat("Combined condition counts:\n")
print(combined_counts)
```

Merge ordered vectors preserving order. This function was created to combine survey data from slightly varied versions:

```{r list-ops-merge}
site1_conditions <- c("Diabetes", "Hypertension", "Heart Failure", "CKD")
site2_conditions <- c("Hypertension", "COPD", "Diabetes", "Cancer")
site3_conditions <- c("CKD", "Diabetes", "Stroke", "Hypertension")

master_list <- merge_ordered_vectors(list(site1_conditions, site2_conditions, site3_conditions))
cat("\nMaster condition list (ordered by first appearance):\n")
print(master_list)
```

Replace elements based on conditions. This function was created to modify a small portion of elements:

```{r list-ops-replace}
sample_data <- data.frame(
    group = c("A", "B", "A", "C", "B"),
    value = c(1, 2, 3, 4, 5),
    stringsAsFactors = FALSE
)

# Recode groups to descriptive labels
recoded <- replace_elements(
    x = sample_data$group,
    from = c("A", "B", "C"),
    to = c("Control", "Treatment", "Placebo")
)

cat("Original vs recoded:\n")
knitr::kable(data.frame(
    Original = sample_data$group,
    Recoded = recoded
), caption = "Group Recoding")
```

### Data Frame Operations

Fill NA values with last valid value. This function is commonly used to handle merged cells in Excel files - when cells are merged in Excel and read into R, the merged cells become NA and can be filled using this function:

```{r df-ops-fill}
time_series <- c(120, NA, NA, 125, NA, 130, NA, NA)
filled_series <- fill_with_last(time_series)

cat("\nMissing data imputation (LOCF):\n")
knitr::kable(data.frame(
    Original = time_series,
    Filled = filled_series
), caption = "Last Observation Carried Forward")
```

## Complete Data Processing Workflow

Here's a comprehensive workflow combining multiple advanced operations for a realistic clinical research scenario. This workflow demonstrates how to use `screen_data_list()` to filter and join multiple tables based on clinical criteria.

### Scenario: Building an Analysis Cohort for a Multi-Table Diabetes Study

Data sources: Patient demographics, admissions, diagnoses, and labs from different EHR tables.

First, create sample multi-source clinical data:

```{r workflow-setup}
cat("=== Clinical Data Processing Workflow ===\n\n")
cat("Step 1: Creating sample multi-source clinical data...\n")

# Patient demographics table
patient <- data.frame(
    pid = 1:50,
    age = sample(25:75, 50, replace = TRUE),
    gender = sample(c("M", "F"), 50, replace = TRUE),
    stringsAsFactors = FALSE
)

# Admission records
admission <- data.frame(
    pid = c(1, 1, 2, 2, 3, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
    vid = 101:115,
    admit_day = c(1, 30, 5, 45, 10, 60, 15, 20, 25, 35, 40, 50, 55, 65, 70),
    stringsAsFactors = FALSE
)

# Diagnosis records
diagnosis <- data.frame(
    pid = c(1, 1, 2, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
    vid = c(101, 102, 103, 104, 105, 107, 108, 109, 110, 111, 112, 113, 114, 115),
    dx_day = c(1, 30, 5, 45, 10, 15, 20, 25, 35, 40, 50, 55, 65, 70),
    icd = c("E11", "I10", "E11", "N18", "E11", "I25", "E11", "J18", "E11", "I10", "E11", "N18", "E11", "I50"),
    stringsAsFactors = FALSE
)

# Lab results
lab <- data.frame(
    pid = c(1, 1, 2, 2, 3, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
    vid = c(101, 102, 103, 104, 105, 106, 107, 108, 109, 110, 111, 112, 113, 114, 115),
    lab_day = c(1, 30, 5, 45, 10, 60, 15, 20, 25, 35, 40, 50, 55, 65, 70),
    HbA1c = c(8.5, 7.2, 9.1, 7.8, 8.0, 6.9, 7.5, 8.2, 7.0, 9.5, 7.8, 8.1, 7.3, 8.8, 7.6),
    stringsAsFactors = FALSE
)

cat("Data sources created:\n")
cat("- Patients:", nrow(patient), "patients\n")
cat("- Admissions:", nrow(admission), "records\n")
cat("- Diagnoses:", nrow(diagnosis), "records\n")
cat("- Labs:", nrow(lab), "results\n")
```

### Step 2: Screen Data Using Clinical Criteria

Use `screen_data_list()` to filter the multi-table dataset. This example selects patients with diabetes (ICD code E11) and keeps only records from their first diabetes diagnosis onward:

```{r workflow-screen}
cat("\nStep 2: Screening multi-table data with clinical criteria...\n")

screened_data <- screen_data_list(
    data_list = list(
        patient = patient,
        admission = admission,
        diagnosis = diagnosis,
        lab = lab
    ),
    # Entry: patients with any diabetes diagnosis
    entry_expr = any(icd == "E11"),
    entry_level = "patient_id",
    # Anchor: keep records from first diabetes diagnosis onward
    anchor_expr = any(icd == "E11"),
    anchor_level = "date",
    anchor_window = "from_first_anchor",
    # Column mappings
    patient_id_map = "pid",
    visit_id_map = c(admission = "vid", diagnosis = "vid", lab = "vid"),
    date_map = c(admission = "admit_day", diagnosis = "dx_day", lab = "lab_day"),
    output = "list",
    return_audit = TRUE
)

cat("Screening complete. Audit log:\n")
knitr::kable(screened_data$audit$entry_scope, caption = "Entry Stage Audit")
knitr::kable(screened_data$audit$anchor_scope, caption = "Anchor Stage Audit")
```

### Step 3: Review Filtered Data

Examine the filtered data for each table:

```{r workflow-review}
cat("\nStep 3: Reviewing filtered data...\n")

cat("\nFiltered patients:")
knitr::kable(head(screened_data$data$patient), caption = "Filtered Patients")

cat("\nFiltered admissions:")
knitr::kable(screened_data$data$admission, caption = "Filtered Admissions")

cat("\nFiltered diagnoses:")
knitr::kable(screened_data$data$diagnosis, caption = "Filtered Diagnoses")

cat("\nFiltered labs:")
knitr::kable(screened_data$data$lab, caption = "Filtered Labs")
```

### Alternative: Joined Output

You can also get the output as a joined data frame instead of a list:

```{r workflow-joined}
cat("\n=== Alternative: Joined Output ===\n")

# For joined output, ensure join keys are unique
# Create a version with unique keys per table
admission_unique <- admission %>%
    mutate(record_id = row_number())

joined_result <- screen_data_list(
    data_list = list(
        patient = patient,
        admission = admission_unique
    ),
    entry_expr = any(age > 40),
    entry_level = "patient_id",
    patient_id_map = "pid",
    output = "joined"
)

cat("Joined data dimensions:", nrow(joined_result), "rows,", ncol(joined_result), "columns\n")
knitr::kable(head(joined_result), caption = "Joined Output")
```

### Workflow Summary

```{r workflow-summary}
cat("\n=== Workflow Summary ===\n")
cat("Initial patients:", nrow(patient), "\n")
cat("Patients with diabetes:", nrow(screened_data$data$patient), "\n")
cat("Retention rate:", round(100 * nrow(screened_data$data$patient) / nrow(patient), 1), "%\n")

cat("\nWorkflow complete!\n")
```

## Summary

The `clinpubr` package provides powerful tools for advanced data operations essential in clinical research:

### Key Functions and Their Applications

| Function | Clinical Application |
|----------|---------------------|
| `exclusion_count()` | CONSORT-style cohort flow documentation |
| `screen_data_list()` | Multi-table cohort screening with clinical criteria |
| `merge_by_substring()` | Substring matching for diagnosis/procedure coding |
| `merge_by_range()` | Time-window matching for longitudinal data |
| `to_wide()` | Format conversion for repeated measures analysis |
| `split_multichoice()` / `combine_multichoice()` | Survey and symptom data processing |
| **Utility functions** | Data standardization and cleaning |

### Best Practices

1. **Document your cohort flow**: Always use `exclusion_count()` to track how your final cohort was derived
2. **Use `screen_data_list()` for complex multi-table studies**: It provides audit trails and handles temporal filtering
3. **Validate substring matching**: When using `merge_by_substring()`, manually review a sample of matches
4. **Check time windows**: Ensure your range matching captures the intended temporal relationships
5. **Handle missing data explicitly**: Don't silently drop missing values - document them
6. **Standardize early**: Apply string standardization and recoding at the beginning of your workflow

### Common Pitfalls to Avoid

- **Over-matching with substrings**: Can create false matches; validate with manual review
- **Overlapping time windows**: Be clear about whether boundaries are inclusive or exclusive
- **Silent data loss**: Always check how many records are lost at each merge step
- **Format confusion**: Document whether your data is in long or wide format at each stage

These functions enable sophisticated data manipulation workflows essential for complex clinical research projects, from retrospective EHR studies to multi-site clinical trials.

## Further Reading

- `vignette("data-cleaning")`: Data cleaning and preparation
- `vignette("baseline-table")`: Creating baseline characteristics tables
- `vignette("regression-analysis")`: Regression analysis and visualization
- `vignette("model-evaluation")`: Model comparison and evaluation
