---
title: "Creating Baseline Characteristics Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Baseline Characteristics Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
library(survival)
```

## Introduction

Baseline characteristics tables are essential in clinical research. They provide a comprehensive summary of patient demographics and clinical features at study entry, allowing readers to assess the comparability of study groups and the generalizability of findings.

### Why Baseline Tables Matter

In clinical trials and observational studies, baseline tables serve several critical purposes:

- **Group comparability**: Assess whether randomization was successful (in RCTs) or whether groups are balanced (in observational studies)
- **Sample description**: Help readers understand the study population
- **Generalizability**: Allow assessment of whether results apply to specific patient populations
- **Statistical planning**: Inform the choice of appropriate statistical methods

### Challenges in Creating Baseline Tables

Creating publication-ready baseline tables involves several challenges:

1. **Variable type determination**: Deciding whether to treat a variable as continuous or categorical
2. **Normality assessment**: Choosing between mean/SD and median/IQR for continuous variables
3. **Statistical test selection**: Selecting appropriate tests (t-test vs Mann-Whitney, Chi-square vs Fisher's exact)
4. **Missing data handling**: Deciding how to report missing values
5. **Multiple comparisons**: Adjusting for multiple testing when comparing more than two groups

The `clinpubr` package automates these decisions while allowing manual override when needed.

## Loading and Preparing Data

We'll use the NCCTG Lung Cancer dataset from the `survival` package, which is commonly used in survival analysis examples.

```{r load-data}
# Load example data - NCCTG Lung Cancer Data
# This dataset contains survival data for patients with advanced lung cancer
# from the North Central Cancer Treatment Group. It's commonly used in 
# survival analysis examples.

data(cancer, package = "survival")

# View the data structure
str(cancer)
knitr::kable(head(cancer), caption = "Raw Data Preview")

# Create some derived variables for demonstration
# Age groups are commonly used in clinical research to stratify analyses
cancer$age_group <- cut(cancer$age, 
                        breaks = c(0, 50, 60, 70, 100),
                        labels = c("<50", "50-60", "60-70", ">70"))

# ECOG performance status is a standard measure of patient functional status
# We combine categories 2 and 3 due to small sample sizes
cancer$ph.ecog_cat <- factor(cancer$ph.ecog,
                             levels = c(0:3),
                             labels = c("0", "1", ">=2", ">=2"))

# Add some missing values for demonstration
# Missing data is common in clinical research due to:
# - Patient refusal to provide certain information
# - Lost to follow-up
# - Data entry errors
set.seed(123)
cancer$meal.cal[sample(1:nrow(cancer), 30)] <- NA
cancer$wt.loss[sample(1:nrow(cancer), 20)] <- NA

# Show data after preparation
knitr::kable(head(cancer[, c("age", "sex", "ph.ecog", "ph.karno", "meal.cal", "wt.loss")]), 
             caption = "Data After Preparation (First 6 Rows)")
```

## Automatic Variable Type Determination

Before creating a baseline table, we need to determine how each variable should be summarized. The `get_var_types()` function automates this process.

### How Variable Types Are Determined

The function analyzes each variable and classifies it as:

- **Factor variables**: Categorical variables (factors or numeric with few unique values)
- **Non-normal variables**: Continuous variables that fail normality tests
- **Exact test variables**: Categorical variables with small cell counts requiring Fisher's exact test
- **Omitted variables**: Variables with too many levels or other issues

```{r get-var-types}
# Automatic variable type determination
# The get_var_types() function analyzes each variable and determines:
# - Whether it's categorical (factor) or continuous
# - For continuous variables: whether it's normally distributed
# - Whether Fisher's exact test is needed (small cell counts)

var_types <- get_var_types(cancer, strata = "sex")

# Display variable classification results
var_summary <- data.frame(
  Category = c("Factor Variables", "Exact Test Variables", "Non-normal Variables", "Omitted Variables"),
  Variables = c(
    paste(var_types$factor_vars, collapse = ", "),
    paste(var_types$exact_vars, collapse = ", "),
    paste(var_types$nonnormal_vars, collapse = ", "),
    paste(var_types$omit_vars, collapse = ", ")
  )
)
knitr::kable(var_summary, caption = "Variable Classification Results")
```

### Customizing Variable Classification

You can adjust the thresholds for automatic classification:

```{r customize-var-types}
# Adjust thresholds for variable classification
# These parameters allow you to customize the automatic classification:
# - num_to_factor: numeric variables with <= this many unique values become factors
# - omit_factor_above: factors with > this many levels are omitted
# - norm_test_by_group: test normality within each stratum

var_types_custom <- get_var_types(
  cancer, 
  strata = "sex",
  num_to_factor = 10,        # Numeric vars with <=10 unique values treated as factor
  omit_factor_above = 15,    # Omit factors with >15 levels
  norm_test_by_group = TRUE  # Test normality within each stratum
)

var_custom_summary <- data.frame(
  Category = c("Factor Variables", "Non-normal Variables"),
  Variables = c(
    paste(var_types_custom$factor_vars, collapse = ", "),
    paste(var_types_custom$nonnormal_vars, collapse = ", ")
  )
)
knitr::kable(var_custom_summary, caption = "Custom Variable Classification")
```

### Visual Inspection with QQ Plots

Sometimes automatic normality tests may not align with visual inspection. You can save QQ plots for manual review:

```{r save-qqplots, eval = FALSE}
# # Save QQ plots for manual review
# # Sometimes automatic normality tests may not align with visual inspection
# # Saving QQ plots allows you to verify the classification
# var_types_with_plots <- get_var_types(
#   cancer,
#   strata = "sex",
#   save_qqplots = TRUE,
#   folder_name = "qqplots_review"
# )
# 
# # The QQ plots are saved in the specified folder for manual inspection
```

## Creating Baseline Tables

### Basic Baseline Table

The `baseline_table()` function automatically:
- Selects appropriate summary statistics (mean/SD vs median/IQR)
- Chooses appropriate statistical tests (t-test vs Mann-Whitney vs Chi-square vs Fisher)
- Calculates p-values for group comparisons

```{r basic-baseline}
# Create baseline table
# The baseline_table() function automatically:
# - Selects appropriate summary statistics (mean/SD vs median/IQR)
# - Chooses appropriate statistical tests (t-test vs Mann-Whitney vs Chi-square vs Fisher)
# - Calculates p-values for group comparisons

baseline_result <- baseline_table(
  cancer,
  var_types = var_types,
  save_table = FALSE
)

# View the baseline table
knitr::kable(baseline_result$baseline, caption = "Baseline Characteristics by Sex")
```

### Multi-Group Comparisons

When there are more than 2 groups, pairwise comparisons are automatically generated. This is useful for studies with multiple treatment arms or disease stages.

```{r multi-group}
# Create baseline table with 3+ groups
# When there are more than 2 groups, pairwise comparisons are automatically generated
# This is useful for studies with multiple treatment arms or disease stages

data(cancer, package = "survival")
cancer$ph.ecog_cat <- factor(cancer$ph.ecog,
                             levels = c(0:3),
                             labels = c("0", "1", ">=2", ">=2"))

var_types_ecog <- get_var_types(cancer, strata = "ph.ecog_cat")

baseline_multi <- baseline_table(
  cancer,
  var_types = var_types_ecog,
  save_table = FALSE,
  multiple_comparison_test = TRUE,
  p_adjust_method = "BH"  # Benjamini-Hochberg adjustment for multiple comparisons
)

# View main baseline table
knitr::kable(baseline_multi$baseline, caption = "Baseline Characteristics by ECOG Status")

# View pairwise comparison results
knitr::kable(baseline_multi$pairwise, caption = "Pairwise Comparison P-values")
```

### Customizing the Baseline Table

You can customize various aspects of the table:

```{r customize-baseline}
# Customize the baseline table output
# You can:
# - Select specific variables to include
# - Add standardized mean differences (SMD) - useful for propensity score studies
# - Handle missing strata values
# - Set seed for reproducible Fisher's exact test simulations

baseline_custom <- baseline_table(
  cancer,
  var_types = var_types,
  vars = c("age", "wt.loss", "meal.cal", "ph.ecog"),  # Select specific variables
  smd = TRUE,                                          # Include standardized mean difference
  omit_missing_strata = TRUE,                          # Remove rows with missing strata
  seed = 123                                           # For reproducible Fisher's exact test
)

knitr::kable(baseline_custom$baseline, caption = "Customized Baseline Table")
```

### Missing Data Summary

The missing data table shows the pattern of missing data by group. This is important for assessing potential bias from missing data.

```{r missing-table}
# The missing data table is included in the results
# This shows the pattern of missing data by group
# Important for assessing potential bias from missing data

knitr::kable(baseline_result$missing, caption = "Missing Data Summary")
```

## Normality Testing

### Distribution Characteristics

Understanding the distribution of your continuous variables helps determine the appropriate summary statistics:

```{r normality-testing}
# Test normality for a single variable
# The test_normality() function uses multiple tests:
# - Shapiro-Wilk test
# - Lilliefors test  
# - Anderson-Darling test
# - Jarque-Bera test
# - Shapiro-Francia test

# Check distribution characteristics using basic statistics
age_data <- cancer$age
wtloss_data <- cancer$wt.loss

# Simple skewness calculation (without moments package)
calc_skewness <- function(x) {
  x <- na.omit(x)
  n <- length(x)
  mean_x <- mean(x)
  sd_x <- sd(x)
  sum((x - mean_x)^3) / ((n - 1) * sd_x^3)
}

# Check distribution characteristics
norm_check <- data.frame(
  Variable = c("age", "wt.loss"),
  Mean = c(mean(age_data, na.rm = TRUE), mean(wtloss_data, na.rm = TRUE)),
  Median = c(median(age_data, na.rm = TRUE), median(wtloss_data, na.rm = TRUE)),
  SD = c(sd(age_data, na.rm = TRUE), sd(wtloss_data, na.rm = TRUE)),
  Skewness = c(
    calc_skewness(age_data),
    calc_skewness(wtloss_data)
  )
)
knitr::kable(norm_check, caption = "Distribution Characteristics")
```

### Adaptive Alpha for Large Samples

With large samples, normality tests become too sensitive and may reject normality even for approximately normal distributions. The package uses an adaptive alpha that decreases with sample size:

```{r adaptive-alpha}
# With small samples (n < 100), alpha = 0.05
# With larger samples, alpha decreases according to an empirical formula
# This addresses the problem that normality tests become too sensitive
# with large sample sizes

# The alpha adjustment formula used in clinpubr:
# For n < 100: alpha = 0.05
# For n >= 100: alpha = 10^(-exp((log10(n) - 2) * 2 + log(-log10(0.05))))

alpha_by_n <- function(n) {
  if (n < 100) {
    return(0.05)
  } else {
    return(10^(-exp((log10(n) - 2) * 2 + log(-log10(0.05)))))
  }
}

alpha_demo <- data.frame(
  Sample_Size = c(50, 100, 500, 1000, 5000),
  Alpha_Used = sapply(c(50, 100, 500, 1000, 5000), alpha_by_n)
)
knitr::kable(alpha_demo, caption = "Adaptive Alpha by Sample Size")
```

## Manual Override

Sometimes you may want to override automatic classification based on clinical knowledge:

```{r manual-override}
# Manually specify variable types
# Sometimes you may want to override automatic classification
# For example, when clinical knowledge suggests a different approach

baseline_manual <- baseline_table(
  cancer,
  strata = "sex",
  factor_vars = c("ph.ecog", "pat.karno"),  # Force these as factors
  nonnormal_vars = c("age"),                 # Force age as non-normal
  exact_vars = c("ph.ecog")                  # Force Fisher's exact test
)

knitr::kable(baseline_manual$baseline, caption = "Baseline Table with Manual Overrides")
```

## Saving Results

You can save all tables to files for further editing or submission:

```{r save-results, eval = FALSE}
# # Save all tables to files
# baseline_saved <- baseline_table(
#   cancer,
#   var_types = var_types,
#   save_table = TRUE,
#   filename = "baseline_characteristics.csv"
# )
# 
# # This creates three files:
# # - baseline_characteristics.csv (main table)
# # - baseline_characteristics_missing.csv (missing data table)
# # - baseline_characteristics_pairwise.csv (pairwise comparisons, if applicable)
```

## Complete Workflow

Here's a complete workflow from data preparation to final table:

```{r complete-workflow}
# Complete workflow from data preparation to final table
# This demonstrates best practices for creating publication-ready tables

# Step 1: Prepare your data
data(cancer, package = "survival")

# Clean and prepare variables with clear, descriptive labels
cancer_clean <- cancer %>%
  mutate(
    # Create categorical variables with descriptive labels
    age_group = cut(age, breaks = c(0, 50, 60, 70, 100), 
                    labels = c("<50", "50-60", "60-70", ">70")),
    
    # Recode ECOG performance status - combine rare categories
    ph.ecog_cat = factor(ph.ecog,
                         levels = c(0:3),
                         labels = c("0", "1", ">=2", ">=2")),
    
    # Ensure proper factor levels for sex with descriptive labels
    sex = factor(sex, labels = c("Male", "Female"))
  )

# Step 2: Determine variable types
# This step is crucial for ensuring appropriate statistical methods
var_types <- get_var_types(
  cancer_clean,
  strata = "sex",
  num_to_factor = 5  # Be conservative with factor classification
)

# Step 3: Review variable classification (optional but recommended)
# This helps catch any misclassifications before generating the final table
var_review <- data.frame(
  Variable_Type = c("Factor", "Non-normal", "Exact"),
  Variables = c(
    paste(var_types$factor_vars, collapse = ", "),
    paste(var_types$nonnormal_vars, collapse = ", "),
    paste(var_types$exact_vars, collapse = ", ")
  )
)
knitr::kable(var_review, caption = "Variable Type Review")

# Step 4: Create baseline table with all recommended options
baseline_final <- baseline_table(
  cancer_clean,
  var_types = var_types,
  smd = TRUE,                      # Include SMD for balance assessment
  multiple_comparison_test = TRUE  # Enable pairwise comparisons if applicable
)

# Step 5: Review results
knitr::kable(baseline_final$baseline, caption = "Final Baseline Characteristics Table")
knitr::kable(baseline_final$missing, caption = "Final Missing Data Summary")
```

## Summary

This vignette covered the essential functions for creating baseline characteristics tables in `clinpubr`:

### Key Functions

- **`get_var_types()`**: Automatic variable type determination
- **`baseline_table()`**: Create comprehensive baseline tables
- **`test_normality()`**: Assess distribution normality

### Best Practices

1. **Always review automatic classifications**: The automatic variable type determination is helpful but should be reviewed
2. **Consider clinical context**: Sometimes clinical knowledge should override statistical classifications
3. **Handle missing data transparently**: Report missing data patterns in your tables
4. **Use appropriate tests**: Let the package choose tests automatically, but verify they make sense
5. **Include SMD for observational studies**: Standardized mean differences help assess group balance

### Interpretation Guidelines

- **P-values**: Values < 0.05 suggest statistically significant differences between groups
- **SMD**: Values > 0.1 indicate potentially meaningful imbalance (commonly used threshold)
- **Missing data**: High missing rates in specific groups may indicate selection bias

These tools help create publication-ready baseline tables that meet the standards of top medical journals.

## Further Reading

- `vignette("data-cleaning")`: Data cleaning and preparation
- `vignette("regression-analysis")`: Regression analysis and visualization
- `vignette("model-evaluation")`: Model comparison and evaluation
