---
title: "Data Cleaning and Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Cleaning and Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
```

## Introduction

Clinical data from Electronic Health Records (EHR) often contains inconsistencies, missing values, outliers, and formatting issues that must be addressed before analysis. This vignette demonstrates a structured cleaning workflow using `clinpubr`:

1. **Data Overview** --- Assess data quality
2. **Format Cleaning** --- Extract numbers, convert dates, check non-numeric values
3. **Unit Standardization** --- Harmonize measurement units
4. **Missing Value Handling** --- Filter or impute missing data
5. **Outlier Detection** --- Identify and manage outliers
6. **Data Transformation** --- Create derived variables

> **Why this order?** Format cleaning and unit standardization must precede outlier detection --- mixed units (e.g., mg/dL vs mmol/L) can cause false outlier flags.

## Creating Realistic Messy Data

Let's create a sample dataset mimicking real-world EHR data with common quality issues:

```{r create-messy-data}
set.seed(123)
n <- 100

# Age with unrealistic values (999 often means "unknown" in legacy systems)
age_vals <- c(sample(18:85, 95, replace = TRUE), 999, 150, -5, 200, 999)
age_vals <- age_vals[1:n]

# BMI with missing-value codes and NAs
bmi_vals <- c(rnorm(90, mean = 24, sd = 4), rep(999, 5), rep(NA, 5))
bmi_vals <- bmi_vals[1:n]

# Gender with inconsistent coding across departments
gender_vals <- sample(c("M", "F", "Male", "Female", "m", "f", "1", "0", NA),
                      n, replace = TRUE)

# Blood pressure stored as character with various formats
systolic_bp_vals <- sample(c(
  paste0(sample(100:160, 80, replace = TRUE), " mmHg"),
  as.character(sample(100:160, 15, replace = TRUE)),
  rep("unknown", 5)
), n, replace = TRUE)

# Visit date with various formats and impossible dates
visit_date_vals <- sample(c(
  format(Sys.Date() - sample(1:365, 60, replace = TRUE), "%Y-%m-%d"),
  format(Sys.Date() - sample(1:365, 30, replace = TRUE), "%Y/%m/%d"),
  "1900-01-01", "2030-12-31", "N/A", "", rep(NA, 6)
), n, replace = TRUE)

# Creatinine: mix of mg/dL and umol/L (1 mg/dL = 88.4 umol/L)
creatinine_vals <- c(
  round(rnorm(80, mean = 1.0, sd = 0.2), 2),
  round(rnorm(15, mean = 88, sd = 15), 2),
  rep(NA, 5)
)
creatinine_vals <- creatinine_vals[1:n]

# HbA1c with text entries and European decimal commas
hba1c_vals <- c(
  round(rnorm(85, mean = 6.5, sd = 1.2), 2),
  rep("<4.0", 3), rep(">14", 2),
  "6,5", "7,2", "5..21", "8..3",
  rep(NA, 6)
)
hba1c_vals <- hba1c_vals[1:n]

messy_data <- data.frame(
  patient_id = 1:n,
  age = age_vals, bmi = bmi_vals, gender = gender_vals,
  systolic_bp = systolic_bp_vals, visit_date = visit_date_vals,
  creatinine = creatinine_vals, hba1c = hba1c_vals,
  stringsAsFactors = FALSE
)

knitr::kable(head(messy_data[sample(n), ], 10),
             caption = "Original Messy Data (10 random rows)")
```

## Step 1: Data Overview

Use `data_overview()` to get a comprehensive diagnostic report of data quality issues:

```{r data-overview}
overview <- data_overview(messy_data)

print(overview$variable_types)
print(overview$summary_stats)
print(overview$quality_issues$numeric_as_character$hba1c)
```

## Step 2: Format Cleaning

### Extracting Numeric Values

Extract clean numbers from free-text fields such as blood pressure recorded with units:

```{r extract-num}
systolic_clean <- extract_num(messy_data$systolic_bp)

knitr::kable(head(data.frame(
  original = messy_data$systolic_bp,
  cleaned = systolic_clean
)), caption = "Blood Pressure: Before vs After Cleaning")
```

### Checking for Non-numeric Values

Identify non-numeric entries before converting to numeric:

```{r check-nonnum}
nonnum_df <- df_view_nonnum(messy_data)
knitr::kable(head(nonnum_df), caption = "Non-numeric Entries by Variable")
```

### Date Conversion

Convert various date formats of `string` type to standard `Date` objects:

```{r date-conversion}
clean_data <- messy_data
clean_data$visit_date_clean <- to_date(messy_data$visit_date)

knitr::kable(head(data.frame(
  original = messy_data$visit_date,
  cleaned = clean_data$visit_date_clean
)), caption = "Visit Dates: Before vs After Conversion")
```

## Step 3: Unit Standardization

Unit standardization should be performed **before** outlier detection. This is critical when merging data from different laboratories:

```{r unit-standardization}
lab_data <- data.frame(
  subject = rep(c("Glucose", "Creatinine", "Cholesterol"), each = 10),
  value = c(
    rnorm(5, 100, 15), rnorm(5, 5.5, 0.8),
    rnorm(5, 1.0, sd = 0.2), rnorm(5, 90, 15),
    rnorm(5, 180, 30), rnorm(5, 4.5, 0.8)
  ),
  unit = c(
    rep("mg/dL", 5), rep("mmol/L", 5),
    rep("mg/dL", 5), rep("umol/L", 5),
    rep("mg/dL", 5), rep("mmol/L", 5)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(unit_view(lab_data, subject_col = "subject", value_col = "value", unit_col = "unit"),
  caption = "Unit Conflicts by Test Type")
```

Standardize using conversion factors (Glucose: mmol/L x 18 = mg/dL; Creatinine: umol/L / 88.4 = mg/dL; Cholesterol: mmol/L x 38.67 = mg/dL):

```{r unit-standardize-apply}
change_rules <- list(
  list(subject = "Glucose", target_unit = "mg/dL", units2change = "mmol/L", coeffs = 18),
  list(subject = "Creatinine", target_unit = "mg/dL", units2change = "umol/L", coeffs = 1 / 88.4),
  list(subject = "Cholesterol", target_unit = "mg/dL", units2change = "mmol/L", coeffs = 38.67)
)

lab_data_std <- unit_standardize(
  lab_data, subject_col = "subject", value_col = "value",
  unit_col = "unit", change_rules = change_rules
)

knitr::kable(head(lab_data_std, 15), caption = "Lab Data After Unit Standardization")
```

## Step 4: Missing Value Handling

For missing data filtering by row/column missing rate, use `get_valid_subset()` to obtain a complete-enough subset for analysis; for more sophisticated imputation (e.g., multiple imputation by chained equations), we recommend the [**mice**](https://cran.r-project.org/package=mice) package.

```{r missing-value}
clean_data <- get_valid_subset(
  messy_data,
  row_na_ratio = 0.5,
  col_na_ratio = 0.02
)
cat("Rows:", nrow(messy_data), "->", nrow(clean_data), "\n")
```

## Step 5: Outlier Detection

> **Important**: Perform outlier detection **after** format cleaning and unit standardization.

The package provides three methods, each with different strengths:

- **MAD** (Median Absolute Deviation): Robust to outliers, good for small samples
- **IQR** (Interquartile Range): Most commonly used in clinical research
- **Z-score**: Assumes normal distribution, use with caution for skewed data

```{r outlier-detection}
# Detect outliers using IQR method
outlier_result <- detect_outliers(clean_data$bmi, method = "iqr")

# Compare methods
knitr::kable(data.frame(
  Method = c("MAD", "IQR", "Z-score"),
  Outlier_Count = c(
    sum(mad_outlier(clean_data$bmi, threshold = 3), na.rm = TRUE),
    sum(iqr_outlier(clean_data$bmi, threshold = 1.5), na.rm = TRUE),
    sum(zscore_outlier(clean_data$bmi, threshold = 3), na.rm = TRUE)
  )
), caption = "Outlier Detection Results by Method")
```

Create indicator variables for outliers to support sensitivity analyses:

```{r mark-outliers}
clean_data$bmi_outlier <- detect_outliers(clean_data$bmi, method = "iqr")$outlier_mask

knitr::kable(data.frame(
  Outlier_Status = c("Not Outlier", "Outlier", "Missing"),
  Count = c(
    sum(clean_data$bmi_outlier == FALSE, na.rm = TRUE),
    sum(clean_data$bmi_outlier == TRUE, na.rm = TRUE),
    sum(is.na(clean_data$bmi_outlier))
  )
), caption = "BMI Outlier Summary")
```

## Step 6: Data Transformation

### Cutting Numeric Variables

Create categorical variables from continuous ones using clinical cut-points:

```{r cut-by}
clean_data$age_clean <- ifelse(messy_data$age > 100 | messy_data$age < 0, NA, messy_data$age)

clean_data$age_group <- cut_by(
  clean_data$age_clean,
  breaks = c(40, 60, 80),
  labels = c("<40", "40-60", "60-80", ">80"),
  label_with_range = TRUE
)

knitr::kable(data.frame(
  Age_Group = names(table(clean_data$age_group)),
  Count = as.vector(table(clean_data$age_group))
), caption = "Age Group Distribution")
```

Cut by quantiles for balanced groups:

```{r cut-by-quantile}
clean_data$bmi_clean <- ifelse(messy_data$bmi == 999, NA, messy_data$bmi)
clean_data$bmi_quartile <- cut_by(
  clean_data$bmi_clean,
  breaks = c(0.25, 0.5, 0.75),
  breaks_as_quantiles = TRUE
)

knitr::kable(data.frame(
  BMI_Quartile = names(table(clean_data$bmi_quartile)),
  Count = as.vector(table(clean_data$bmi_quartile))
), caption = "BMI Quartile Distribution")
```

## Complete Cleaning Pipeline

Here's a complete workflow from raw data to analysis-ready data:

```{r complete-pipeline}
# Step 1: Overview
overview <- data_overview(messy_data)

# Step 2: Format cleaning
clean <- messy_data
clean$systolic_bp <- extract_num(clean$systolic_bp)
clean$visit_date <- to_date(clean$visit_date)

# Step 3: Missing value filtering
clean <- get_valid_subset(clean, row_na_ratio = 0.5, col_na_ratio = 0.3)

# Step 4: Outlier handling
outlier_res <- detect_outliers(clean$bmi, method = "iqr")
clean$bmi <- ifelse(outlier_res$outlier_mask, NA, clean$bmi)

# Step 5: Derived variables
clean$age_group <- cut_by(
  clean$age, breaks = c(40, 60, 80),
  labels = c("<40", "40-60", "60-80", ">80")
)

# Step 6: Final check
final_overview <- data_overview(clean)
knitr::kable(final_overview, caption = "Final Data Quality Overview")

cat("\nOriginal rows:", nrow(messy_data),
    "| Final rows:", nrow(clean_data),
    "| Removed:", nrow(messy_data) - nrow(clean_data), "\n")
```

## Summary

| Step | Function | Purpose |
|------|----------|---------|
| 1. Data Overview | `data_overview()` | Comprehensive data quality assessment |
| 2. Format Cleaning | `extract_num()`, `to_date()`, `check_nonnum()` | Fix formatting issues |
| 3. Unit Standardization | `unit_standardize()` | Standardize measurement units |
| 4. Missing Values | `get_valid_subset()` | Filter by missing rate |
| 5. Outlier Detection | `detect_outliers()`, `mad_outlier()`, `iqr_outlier()` | Detect and manage outliers |
| 6. Data Transformation | `cut_by()` | Create derived variables |

## Further Reading

- `vignette("baseline-table")`: Creating publication-ready baseline tables
- `vignette("regression-analysis")`: Regression analysis and visualization
- `vignette("model-evaluation")`: Model comparison and evaluation
