---
title: "Data Cleaning and Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Cleaning and Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
```

## Introduction

Clinical data cleaning is a critical step in medical research. Electronic Health Records (EHR) often contain inconsistencies, missing values, outliers, and formatting issues that need to be addressed before analysis. The `clinpubr` package provides a comprehensive set of tools to streamline this process.

### Real-World Data Challenges in Clinical Research

In clinical practice, data quality issues are common due to:

- **System migrations**: Legacy systems may use different coding schemes (e.g., 999 for "unknown")
- **Manual entry errors**: Typos, impossible values (negative age, future dates)
- **Inconsistent formats**: Different departments using different date formats or units
- **Free-text entries**: Blood pressure recorded as "120 mmHg", "120", or "unknown"
- **Unit variations**: Lab results in mg/dL vs mmol/L depending on the laboratory
- **Semantic variations**: Different names for the same entity (e.g., "creatinine", "CREA", "Scr" all referring to serum creatinine)

### Data Cleaning Workflow

This vignette follows a structured data cleaning workflow:

1. **Data Overview** — Get a quick overview of your data quality
2. **Format Cleaning** — Fix formatting issues (extract numbers, convert dates, check non-numeric values)
3. **Unit Standardization** — Standardize measurement units across sources
4. **Missing Value Handling** — Understand and handle missing data patterns
5. **Outlier Detection** — Detect and manage outliers
6. **Data Transformation** — Create derived variables (grouping, multi-choice splitting)
7. **Complete Pipeline** — Put it all together

> **Why this order?** Format cleaning and unit standardization must come before outlier detection — mixed units (e.g., mg/dL vs mmol/L) can cause false outlier flags. Missing value handling and outlier detection should follow once the data is in a clean, consistent format.

## Creating Realistic Messy Data

To demonstrate the data cleaning workflow, let's create a sample dataset that mimics real-world EHR data with common quality issues. In clinical practice, we often encounter data entry errors, inconsistent formats, and missing values due to various reasons: system migrations, manual entry mistakes, or different data collection protocols across departments.

```{r create-messy-data}
set.seed(123)
n <- 100

# Age with some unrealistic values (999 often means "unknown" in legacy systems)
age_vals <- c(sample(18:85, 95, replace = TRUE), 999, 150, -5, 200, 999)
age_vals <- age_vals[1:n] # Ensure exactly n values

# BMI with outliers and missing values
# In real EHR data, 999 or -99 are commonly used as missing value codes
bmi_vals <- c(rnorm(90, mean = 24, sd = 4), rep(999, 5), rep(NA, 5))
bmi_vals <- bmi_vals[1:n]

# Gender with inconsistent coding
# Different departments may use different coding schemes
gender_vals <- sample(c("M", "F", "Male", "Female", "m", "f", "1", "0", NA), n, replace = TRUE)

# Blood pressure stored as character with various formats
# This often happens when data is copied from free-text notes
systolic_bp_vals <- sample(c(
  paste0(sample(100:160, 80, replace = TRUE), " mmHg"),
  as.character(sample(100:160, 15, replace = TRUE)),
  rep("unknown", 5)
), n, replace = TRUE)

# Visit date with various formats and impossible dates
# Legacy systems may have default dates like 1900-01-01 or future dates
visit_date_vals <- sample(c(
  format(Sys.Date() - sample(1:365, 60, replace = TRUE), "%Y-%m-%d"),
  format(Sys.Date() - sample(1:365, 30, replace = TRUE), "%Y/%m/%d"),
  "1900-01-01", "2030-12-31", "N/A", "", rep(NA, 6)
), n, replace = TRUE)

# Lab values with different units
# Creatinine is particularly problematic: mg/dL vs umol/L (1 mg/dL = 88.4 umol/L)
creatinine_vals <- c(
  round(rnorm(80, mean = 1.0, sd = 0.2), 2), # mg/dL
  round(rnorm(15, mean = 88, sd = 15), 2), # umol/L - different unit!
  rep(NA, 5)
)
creatinine_vals <- creatinine_vals[1:n]

# HbA1c with text entries and different decimal separators
hba1c_vals <- c(
  round(rnorm(85, mean = 6.5, sd = 1.2), 2),
  rep("<4.0", 3), # Below detection limit
  rep(">14", 2), # Above detection limit
  "6,5", "7,2", "5..21", "8..3", # European decimal comma and extra dot
  rep(NA, 6)
)
hba1c_vals <- hba1c_vals[1:n]

# Create the data frame
messy_data <- data.frame(
  patient_id = 1:n,
  age = age_vals,
  bmi = bmi_vals,
  gender = gender_vals,
  systolic_bp = systolic_bp_vals,
  visit_date = visit_date_vals,
  creatinine = creatinine_vals,
  hba1c = hba1c_vals,
  stringsAsFactors = FALSE
)

knitr::kable(head(messy_data[sample(n), ], 10), caption = "Original Messy Data (10 random Rows)")
```

Notice the various data quality issues in this sample:

- **Age**: Contains impossible values like -5, 999, 150, 200
- **BMI**: Has 999 as a missing code and some NA values
- **Gender**: Multiple coding schemes (M/F, Male/Female, 1/0, different cases)
- **Blood Pressure**: Mixed formats with units and plain numbers
- **Visit Date**: Multiple date formats and impossible dates (1900, 2030)
- **Creatinine**: Mix of mg/dL and umol/L values
- **HbA1c**: Contains text entries and European decimal commas

## Step 1: Data Overview

The first step in data cleaning is understanding what issues exist in your dataset. The `data_overview()` function provides a comprehensive diagnostic report.

```{r data-overview}
overview <- data_overview(messy_data)

# the `overview` object stores the details in different components, for example:
print(overview$variable_types)
print(overview$summary_stats)
print(overview$quality_issues$numeric_as_character$hba1c)
```

The `data_overview()` function identifies the following **data quality issues** and provides recommendations:

- **Missing data patterns**: Which variables have missing values
- **Potential outliers**: Values that seem unrealistic
- **Suspicious values**: Values that may indicate data entry errors or datatype issues
- **Inconsistent formats**: Mixed formats within the same variable

## Step 2: Format Cleaning

Before performing any statistical operations, we need to fix basic formatting issues — extracting numbers from text, converting dates, and identifying non-numeric entries.

### Extracting Numeric Values

Extract and clean numeric values from character strings. This is common when dealing with free-text entries such as blood pressure recorded with units.

Before extraction: mixed formats with units and text:

```{r extract-num-before}
knitr::kable(head(data.frame(
  original = messy_data$systolic_bp
)), caption = "Blood Pressure Before Cleaning")
```

After extraction: clean numeric values:

```{r extract-num-after}
systolic_clean <- extract_num(messy_data$systolic_bp)

knitr::kable(head(data.frame(
  original = messy_data$systolic_bp,
  cleaned = systolic_clean
)), caption = "Blood Pressure: Before vs After Cleaning")
```

### Checking for Non-numeric Values

For variables that should be numeric but contain non-numeric entries. This helps identify problematic entries before conversion:

```{r initial-cleaning}
nonnum_values <- check_nonnum(messy_data$systolic_bp)

knitr::kable(data.frame(
  Check = c("Total Non-numeric Values", "Unique Non-numeric Values"),
  Result = c(length(nonnum_values), length(unique(nonnum_values)))
), caption = "Non-numeric Value Check")
```

View all non-numeric entries in the data frame:

```{r view-nonnum}
nonnum_df <- df_view_nonnum(messy_data)
if (nrow(nonnum_df) > 0) {
  knitr::kable(head(nonnum_df), caption = "Non-numeric Entries by Variable")
} else {
  knitr::kable(data.frame(Message = "No non-numeric values found"), caption = "Non-numeric Check Result")
}
```

### Date Conversion

Convert various date formats to standard Date objects. Real-world date data often has multiple formats and invalid dates.

Before conversion:

```{r date-conversion-before}
knitr::kable(head(data.frame(
  original_visit_date = messy_data$visit_date
)), caption = "Visit Dates Before Conversion")
```

After conversion:

```{r date-conversion-after}
clean_data <- messy_data
clean_data$visit_date_clean <- to_date(messy_data$visit_date)

knitr::kable(head(data.frame(
  original = messy_data$visit_date,
  cleaned = clean_data$visit_date_clean,
  is_valid = !is.na(clean_data$visit_date_clean)
)), caption = "Visit Dates: Before vs After Conversion")
```

## Step 3: Unit Standardization

Unit standardization should be performed **before** outlier detection — mixed units (e.g., mg/dL vs mmol/L for the same lab test) can cause valid values to be falsely flagged as outliers.

For clinical laboratory data with mixed units. This is a common problem when merging data from different laboratories or when units change over time in the same institution:

```{r unit-standardization}
lab_data <- data.frame(
  subject = rep(c("Glucose", "Creatinine", "Cholesterol"), each = 10),
  value = c(
    rnorm(5, 100, 15), rnorm(5, 5.5, 0.8), # Glucose: mg/dL and mmol/L
    rnorm(5, 1.0, sd = 0.2), rnorm(5, 90, 15), # Creatinine: mg/dL and umol/L
    rnorm(5, 180, 30), rnorm(5, 4.5, 0.8) # Cholesterol: mg/dL and mmol/L
  ),
  unit = c(
    rep("mg/dL", 5), rep("mmol/L", 5),
    rep("mg/dL", 5), rep("umol/L", 5),
    rep("mg/dL", 5), rep("mmol/L", 5)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(unit_view(lab_data, subject_col = "subject", value_col = "value", unit_col = "unit"),
  caption = "Unit Conflicts by Test Type")
```

Standardize units using conversion factors. These are standard clinical conversion factors:
- Glucose: mmol/L × 18 = mg/dL
- Creatinine: umol/L ÷ 88.4 = mg/dL
- Cholesterol: mmol/L × 38.67 = mg/dL

```{r unit-standardize-apply}
change_rules <- list(
  list(subject = "Glucose", target_unit = "mg/dL", units2change = "mmol/L", coeffs = 18),
  list(subject = "Creatinine", target_unit = "mg/dL", units2change = "umol/L", coeffs = 1 / 88.4),
  list(subject = "Cholesterol", target_unit = "mg/dL", units2change = "mmol/L", coeffs = 38.67)
)

lab_data_standardized <- unit_standardize(
  lab_data,
  subject_col = "subject",
  value_col = "value",
  unit_col = "unit",
  change_rules = change_rules
)

knitr::kable(head(lab_data_standardized, 15), caption = "Lab Data After Unit Standardization")
```

## Step 4: Outlier Detection and Handling

> **Important**: Outlier detection should be performed **after** format cleaning and unit standardization. Mixed formats or inconsistent units can cause valid values to be falsely flagged as outliers.

### Detecting Outliers

The package provides multiple methods for outlier detection. Each method has its strengths:

- **MAD (Median Absolute Deviation)**: Robust to outliers, good for small samples
- **IQR (Interquartile Range)**: Most commonly used in clinical research
- **Z-score**: Assumes normal distribution, use with caution for skewed data

MAD is robust to outliers and works well for small samples:

```{r outlier-detection-mad}
mad_outliers <- mad_outlier(messy_data$bmi, threshold = 3)
mad_count <- sum(mad_outliers, na.rm = TRUE)
```

IQR is the most commonly used method in clinical research:

```{r outlier-detection-iqr}
iqr_outliers <- iqr_outlier(messy_data$bmi, threshold = 1.5)
iqr_count <- sum(iqr_outliers, na.rm = TRUE)
```

Z-score assumes normal distribution - use with caution for skewed data:

```{r outlier-detection-zscore}
zscore_outliers <- zscore_outlier(messy_data$bmi, threshold = 3)
zscore_count <- sum(zscore_outliers, na.rm = TRUE)
```

Comprehensive outlier detection with detailed output:

```{r outlier-detection-summary}
outlier_result <- detect_outliers(messy_data$bmi, method = "mad")

knitr::kable(data.frame(
  Method = c("MAD", "IQR", "Z-score"),
  Outlier_Count = c(mad_count, iqr_count, zscore_count)
), caption = "Outlier Detection Results by Method")
```

### Marking Outliers

You can create indicator variables for outliers. This allows you to:

1. Exclude outliers in sensitivity analyses
2. Investigate why certain values are outliers
3. Report how many outliers were handled

```{r mark-outliers}
clean_data <- messy_data
outlier_result <- detect_outliers(clean_data$bmi, method = "iqr")
clean_data$bmi_outlier <- outlier_result$outlier_mask

knitr::kable(data.frame(
  Outlier_Status = c("Not Outlier", "Outlier", "Missing"),
  Count = c(
    sum(clean_data$bmi_outlier == FALSE, na.rm = TRUE),
    sum(clean_data$bmi_outlier == TRUE, na.rm = TRUE),
    sum(is.na(clean_data$bmi_outlier))
  )
), caption = "BMI Outlier Summary")
```

## Step 5: Data Transformation

### Cutting Numeric Variables

Create categorical variables from continuous ones. This is useful for:

- Creating clinically meaningful groups (e.g., age groups)
- Stratifying analyses
- Creating balanced groups using quantiles

Cut age into clinically meaningful groups:

```{r cut-by-age}
clean_data$age <- messy_data$age # Keep original for comparison
clean_data$age_clean <- ifelse(messy_data$age > 100 | messy_data$age < 0, NA, messy_data$age)

clean_data$age_group <- cut_by(
  clean_data$age_clean,
  breaks = c(40, 60, 80),
  labels = c("<40", "40-60", "60-80", ">80"),
  label_with_range = TRUE
)

knitr::kable(data.frame(
  Age_Group = names(table(clean_data$age_group)),
  Count = as.vector(table(clean_data$age_group))
), caption = "Age Group Distribution")
```

Cut by quantiles (useful for creating balanced groups):

```{r cut-by-quantile}
clean_data$bmi_clean <- ifelse(messy_data$bmi == 999, NA, messy_data$bmi)
clean_data$bmi_quartile <- cut_by(
  clean_data$bmi_clean,
  breaks = c(0.25, 0.5, 0.75),
  breaks_as_quantiles = TRUE
)

knitr::kable(data.frame(
  BMI_Quartile = names(table(clean_data$bmi_quartile)),
  Count = as.vector(table(clean_data$bmi_quartile))
), caption = "BMI Quartile Distribution")
```

## Complete Cleaning Pipeline

Here's a complete workflow for cleaning clinical data — from raw data to analysis-ready data. Notice the order: **format cleaning → unit standardization → missing values → outliers → derived variables**.

Step 1: Get data overview to identify issues:

```{r complete-pipeline-step1}
cat("Step 1: Data Overview\n")
overview <- data_overview(messy_data)
```

Step 2: Format cleaning — extract numbers from text fields and convert dates:

```{r complete-pipeline-step2}
cat("Step 2: Format Cleaning\n")
clean_data <- messy_data
clean_data$systolic_bp <- extract_num(clean_data$systolic_bp)
clean_data$visit_date <- to_date(clean_data$visit_date)
```

Step 3: Handle missing values — get valid subset. Adjust thresholds based on your analysis requirements:

```{r complete-pipeline-step3}
cat("Step 3: Handling Missing Values\n")
clean_data <- get_valid_subset(
  clean_data,
  row_na_ratio = 0.5, # More lenient for this example
  col_na_ratio = 0.3
)
```

Step 4: Handle outliers — replace outliers with NA for further investigation:

```{r complete-pipeline-step4}
cat("Step 4: Handling Outliers\n")
outlier_result_final <- detect_outliers(clean_data$bmi, method = "iqr")
clean_data$bmi <- ifelse(
  outlier_result_final$outlier_mask,
  NA, # Mark outliers as NA for review
  clean_data$bmi
)
```

Step 5: Create derived variables:

```{r complete-pipeline-step5}
cat("Step 5: Creating Derived Variables\n")
clean_data$age_group <- cut_by(
  clean_data$age,
  breaks = c(40, 60, 80),
  breaks_as_quantiles = FALSE,
  labels = c("<40", "40-60", "60-80", ">80")
)
```

Step 6: Final check:

```{r complete-pipeline-step6}
cat("Step 6: Final Data Quality Check\n")
final_overview <- data_overview(clean_data)
knitr::kable(final_overview, caption = "Final Data Quality Overview")
```

Summary of cleaning results:

```{r complete-pipeline-summary}
cat("\n=== Data Cleaning Summary ===\n")
cat("Original rows:", nrow(messy_data), "\n")
cat("Final rows:", nrow(clean_data), "\n")
cat("Rows removed:", nrow(messy_data) - nrow(clean_data), "\n")
```

## Summary

This vignette covered the essential data cleaning functions in `clinpubr`, following a structured workflow:

| Step | Function | Purpose |
|------|----------|---------|
| 1. Data Overview | `data_overview()` | Comprehensive data quality assessment |
| 2. Format Cleaning | `extract_num()`, `to_date()`, `check_nonnum()` | Fix formatting issues |
| 3. Unit Standardization | `unit_standardize()` | Standardize measurement units |
| 4. Missing Values | `get_valid_subset()`, `get_valid()` | Handle missing data patterns |
| 5. Outlier Detection | `detect_outliers()`, `mad_outlier()`, `iqr_outlier()` | Detect and manage outliers |
| 6. Data Transformation | `cut_by()`, `split_multichoice()` | Create derived variables |

### Best Practices for Clinical Data Cleaning

1. **Always start with data overview**: Understand the scope of data quality issues before cleaning
2. **Clean formats first**: Extract numbers, convert dates — before any statistical operations
3. **Standardize units early**: Do this before outlier detection to avoid false positives
4. **Handle outliers carefully**: Don't automatically remove outliers — investigate them first
5. **Document your cleaning steps**: Keep a record of what was changed and why
6. **Create derived variables last**: Grouping and splitting should follow data cleaning
7. **Final validation**: Run data overview again after cleaning to confirm improvements

These tools form a complete pipeline for preparing clinical data for analysis, ensuring data quality and consistency for subsequent statistical modeling and visualization.

## Further Reading

- `vignette("baseline-table")`: Creating publication-ready baseline tables
- `vignette("regression-analysis")`: Regression analysis and visualization
- `vignette("model-evaluation")`: Model comparison and evaluation
