---
title: "Model Evaluation and Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Evaluation and Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
library(survival)
library(ggplot2)
```

## Introduction

Evaluating clinical prediction models requires assessing discrimination, calibration, and clinical utility. This vignette demonstrates a comprehensive evaluation workflow:

1. **Model Comparison** --- Compare multiple models using discrimination and calibration metrics
2. **Time-Dependent ROC** --- Evaluate survival prediction models at specific time points
3. **C-Index** --- Global discrimination measure for survival models
4. **Variable Importance** --- Identify key predictive features

## Preparing Data

We'll use the NCCTG Lung Cancer dataset and fit three logistic regression models of increasing complexity:

```{r prepare-data}
data(cancer, package = "survival")
cancer$dead <- cancer$status == 2
cancer$event <- ifelse(cancer$status == 2, 1, 0)

cancer_clean <- cancer %>%
  mutate(sex = factor(sex, labels = c("Male", "Female"))) %>%
  na.omit()

# Fit three models with increasing complexity
cancer_clean$pred_model1 <- predict(
  glm(dead ~ age + sex + ph.karno, data = cancer_clean, family = binomial),
  type = "response"
)
cancer_clean$pred_model2 <- predict(
  glm(dead ~ age + sex + ph.karno + pat.karno, data = cancer_clean, family = binomial),
  type = "response"
)
cancer_clean$pred_model3 <- predict(
  glm(dead ~ age + sex + ph.karno + wt.loss, data = cancer_clean, family = binomial),
  type = "response"
)

knitr::kable(head(cancer_clean[, c("dead", "pred_model1", "pred_model2", "pred_model3")]),
             caption = "Model Predictions vs Actual Outcomes (First 6 Rows)")
```

## Model Comparison

`classif_model_compare()` provides a comprehensive comparison using multiple metrics (AUC, Accuracy, Sensitivity, Specificity, PPV, NPV, PRAUC, Brier Score) and generates ROC, PR, calibration, and DCA plots:

```{r model-compare}
model_comparison <- classif_model_compare(
  data = cancer_clean,
  target_var = "dead",
  model_names = c("pred_model1", "pred_model2", "pred_model3"),
  save_output = FALSE
)

knitr::kable(model_comparison$metric_table, caption = "Model Performance Comparison")
```

### Visualization

```{r model-plots-roc}
model_comparison$roc_plot
```

```{r model-plots-pr}
model_comparison$pr_plot
```

```{r model-plots-calibration}
model_comparison$calibration_plot
```

```{r model-plots-dca}
model_comparison$dca_plot
```

## Time-Dependent ROC Analysis

For survival outcomes, time-dependent ROC accounts for the timing of events:

```{r time-roc}
time_roc_result <- time_roc_plot(
  data = cancer_clean,
  event_var = "event",
  time_var = "time",
  marker_var = "pred_model1",
  times = c(200, 365),
  save_plot = FALSE
)

time_roc_result
```

## C-Index

The C-index (concordance index) measures discrimination for survival models. Interpretation: 0.5 = no discrimination, 0.7-0.8 = acceptable, >0.8 = excellent.

```{r c-index}
models <- c("pred_model1", "pred_model2", "pred_model3")
cindex_results <- sapply(models, function(m) {
  calc_cindex(cancer_clean, "time", "event", m)
})

knitr::kable(data.frame(
  Model = c("Basic (Age+Sex+Karnofsky)", "Extended (+Pat.Karnofsky)", "Alternative (+Weight Loss)"),
  C_index = round(cindex_results, 3),
  Rank = rank(-cindex_results)
), caption = "C-index Comparison Across All Models")
```

## Variable Importance

```{r importance-plot}
importance_scores <- c(
  Age = 0.25, Sex = 0.15, Karnofsky = 0.30,
  WeightLoss = 0.12, Calories = 0.08, PatKarnofsky = 0.10
)

importance_plot(x = importance_scores, x_lab = "Variable Importance")
```

Show only top variables with custom styling:

```{r custom-importance}
importance_plot(
  x = importance_scores,
  x_lab = "Relative Importance",
  top_n = 4,
  color = "steelblue"
)
```

## Complete Evaluation Workflow

```{r complete-evaluation}
# Step 1: Data preparation
data(cancer, package = "survival")
eval_data <- cancer %>%
  mutate(
    dead = status == 2,
    event = ifelse(status == 2, 1, 0),
    sex = factor(sex, labels = c("Male", "Female"))
  ) %>%
  na.omit()

# Step 2: Fit models
eval_data$pred_m1 <- predict(glm(dead ~ age + sex, data = eval_data, family = binomial), type = "response")
eval_data$pred_m2 <- predict(glm(dead ~ age + sex + ph.karno, data = eval_data, family = binomial), type = "response")
eval_data$pred_m3 <- predict(glm(dead ~ age + sex + ph.karno + pat.karno + wt.loss,
                                  data = eval_data, family = binomial), type = "response")

# Step 3: Model comparison
comparison <- classif_model_compare(
  eval_data, target_var = "dead",
  model_names = c("pred_m1", "pred_m2", "pred_m3"),
  save_output = FALSE
)

comparison_metrics <- comparison$metric_table
knitr::kable(comparison_metrics, caption = "Comprehensive Model Comparison")

# Step 4: Time-dependent ROC
time_roc <- time_roc_plot(
  eval_data, event_var = "event", time_var = "time",
  marker_var = "pred_m3", times = c(200, 365),
  save_plot = FALSE
)

# Step 5: C-index
cindex_vals <- sapply(c("pred_m1", "pred_m2", "pred_m3"), function(m) {
  calc_cindex(eval_data, "time", "event", m)
})

knitr::kable(data.frame(
  Model = c("Basic (Age+Sex)", "With Performance Status", "Full Model"),
  C_Index = round(cindex_vals, 3)
), caption = "C-index Summary")

# Step 6: Summary
cat("\nBest C-index:", names(which.max(cindex_vals)),
    "(C-index =", round(max(cindex_vals), 3), ")\n")
```

## Summary

### Key Functions

- **`classif_model_compare()`**: Comprehensive model comparison with ROC, PR, calibration, and DCA plots
- **`time_roc_plot()`**: Time-dependent ROC analysis for survival outcomes
- **`calc_cindex()`**: Calculate concordance index for survival models
- **`importance_plot()`**: Visualize variable importance

### Evaluation Checklist

1. **Discrimination**: AUC (binary) or C-index (survival), sensitivity/specificity at clinical thresholds
2. **Calibration**: Calibration plots, Brier score
3. **Clinical utility**: Decision curve analysis, net benefit at relevant thresholds
4. **External validation**: Performance in independent populations

## Further Reading

- `vignette("data-cleaning")`: Data cleaning and preparation
- `vignette("baseline-table")`: Creating baseline characteristics tables
- `vignette("regression-analysis")`: Regression analysis and visualization
