---
title: "Regression Analysis and Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Analysis and Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
library(survival)
library(ggplot2)
```

## Introduction

The `clinpubr` package provides a comprehensive suite for regression analysis in clinical research, supporting Cox proportional hazards, logistic, and linear regression with publication-ready visualizations. This vignette covers the typical analysis workflow:

1. **Regression Scan** --- Screen predictors with multiple transformations
2. **Comprehensive Results** --- Generate multi-model specification tables
3. **Forest Plots** --- Visualize effect sizes
4. **Predictor Effects** --- RCS plots and effect plots
5. **Interaction Analysis** --- Test and visualize effect modification
6. **Subgroup Analysis** --- Subgroup forest plots

## Data Preparation

We'll use the NCCTG Lung Cancer dataset (228 patients with advanced lung cancer):

```{r load-data}
data(cancer, package = "survival")
cancer$dead <- cancer$status == 2

cancer_clean <- cancer %>%
  mutate(
    sex = factor(sex, labels = c("Male", "Female")))
  ) %>%
  na.omit()

knitr::kable(data.frame(
  Rows = nrow(cancer_clean),
  Columns = ncol(cancer_clean),
  Description = "Complete cases for analysis"
), caption = "Analysis Dataset Summary")

knitr::kable(head(cancer_clean), caption = "Sample of Analysis Variables")
```

## Regression Scan

`regression_scan()` automatically tests multiple variable transformations for each predictor:

- **Original** (linear) form
- **Logarithmic** transformation (if all values > 0)
- **Categorical** form (median split)
- **RCS** (Restricted Cubic Spline) for non-linear relationships

### Cox Proportional Hazards Model

```{r cox-scan}
cox_scan <- regression_scan(
  data = cancer_clean,
  y = "status",
  time = "time",
  predictors = c("age", "sex", "ph.karno", "pat.karno", "wt.loss", "meal.cal"),
  covars = NULL,
  save_table = FALSE
)

knitr::kable(cox_scan, caption = "Cox Regression Scan Results")
```

The scan reports the effect size, p-value, adjusted p-value (BH correction by default), and the best transformation.

View significant predictors:

```{r interpret-scan}
significant_predictors <- cox_scan %>%
  filter(original.pval < 0.05 | rcs.overall.pval < 0.05) %>%
  select(predictor, nvalid, best.var.trans, original.pval, rcs.overall.pval)

knitr::kable(significant_predictors,
             caption = "Significant Predictors (p < 0.05)",
             digits = 4)
```

### Logistic Regression Scan

```{r logistic-scan}
logistic_scan <- regression_scan(
  data = cancer_clean,
  y = "dead",
  predictors = c("age", "sex", "ph.karno", "pat.karno", "wt.loss"),
  covars = NULL,
  save_table = FALSE
)

knitr::kable(logistic_scan, caption = "Logistic Regression Scan Results")
```

## Comprehensive Regression Results

In clinical publications, it's standard to present multiple model specifications (Crude, Partially adjusted, Fully adjusted) to show how confounding affects the estimated association.

### Cox Model with Multiple Specifications

```{r cox-basic-results}
cox_results <- regression_basic_results(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  model_covs = list(
    Crude = c(),
    Model1 = c("sex"),
    Model2 = c("sex", "ph.karno", "pat.karno")
  ),
  pers = c(1, 5, 10),
  save_output = FALSE
)

knitr::kable(cox_results$table, caption = "Cox Regression Results for Age")
```

### Logistic Model with Clinical Cut-points

```{r logistic-basic-results}
logistic_results <- regression_basic_results(
  data = cancer_clean,
  x = "ph.karno",
  y = "dead",
  model_covs = list(
    Crude = c(),
    Adjusted = c("age", "sex")
  ),
  factor_breaks = c(80),
  factor_labels = c("<80", ">=80"),
  save_output = FALSE
)

knitr::kable(logistic_results$table, caption = "Logistic Regression Results for Performance Status")
```

### Quantile-Based Categories

When clinical cut-points aren't established, use tertiles for balanced group comparison:

```{r custom-transformations}
custom_results <- regression_basic_results(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  quantile_breaks = c(0.33, 0.67),
  quantile_labels = c("Youngest Third", "Middle Third", "Oldest Third"),
  label_with_range = TRUE,
  save_output = FALSE
)

knitr::kable(custom_results$table, caption = "Cox Regression Results - Age Tertiles")
```

## Forest Plots

### Univariate Forest Plot

```{r univariate-forest}
uni_forest <- regression_forest(
  data = cancer_clean,
  model_vars = c("age", "ph.karno", "pat.karno", "wt.loss"),
  y = "status",
  time = "time",
  as_univariate = TRUE,
  save_plot = FALSE
)

uni_forest
```

The univariate forest plot displays:

- **Left panel**: Variable names with sample sizes in parentheses
- **Center**: Forest plot with hazard ratios (HR) and 95% confidence intervals
  - Each square represents the point estimate (HR)
  - Horizontal lines represent 95% confidence intervals
  - The vertical dashed line at HR=1.0 indicates no effect
  - Squares to the left of 1.0 indicate protective effects (HR < 1)
  - Squares to the right of 1.0 indicate risk-increasing effects (HR > 1)
- **Right panel**: Numerical values of HR (95% CI) and p-values

In this example, `pat.karno` shows a significant protective effect (HR ≈ 0.98, p < 0.01), while `age`, `ph.karno`, and `wt.loss` do not reach statistical significance in univariate analysis.

### Multivariate Forest Plot

```{r multivariate-forest}
multi_forest <- regression_forest(
  data = cancer_clean,
  model_vars = list(
    Model1 = c("age"),
    Model2 = c("age", "ph.karno"),
    Model3 = c("age", "ph.karno", "pat.karno")
  ),
  y = "status",
  time = "time",
  as_univariate = FALSE,
  save_plot = FALSE
)

multi_forest
```

The multivariate forest plot compares different model specifications:

- **Model1**: Contains only `age` (crude model)
- **Model2**: Contains `age` and `ph.karno` (partially adjusted)
- **Model3**: Contains `age`, `ph.karno`, and `pat.karno` (fully adjusted)

Key observations:
- As more covariates are added, the effect estimate for `age` changes slightly
- The confidence intervals may widen or narrow depending on the correlation between predictors
- This visualization helps assess confounding effects and model stability

### Customizing Forest Plots

```{r custom-forest}
custom_forest <- regression_forest(
  data = cancer_clean,
  model_vars = c("age", "ph.karno", "pat.karno", "wt.loss", "meal.cal"),
  y = "status",
  time = "time",
  as_univariate = TRUE,
  show_vars = c("age", "ph.karno", "pat.karno"),
  est_nsmall = 2,
  p_nsmall = 4,
  save_plot = FALSE
)

custom_forest
```

The customized forest plot demonstrates several formatting options:

- **`show_vars`**: Limits displayed variables to `age`, `ph.karno`, and `pat.karno`
- **`est_nsmall = 2`**: Displays effect estimates with 2 decimal places
- **`p_nsmall = 4`**: Displays p-values with 4 decimal places

This allows focused presentation of key predictors with precise numerical formatting suitable for publication.

## Predictor Effect Visualization

### RCS Plots

RCS plots visualize non-linear relationships while preserving the continuous nature of the data:

```{r rcs-plot}
rcs_age <- rcs_plot(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  covars = c("sex", "ph.karno"),
  knot = 4,
  ref = "x_median",
  add_hist = TRUE,
  save_plot = FALSE
)

rcs_age
```

The RCS (Restricted Cubic Spline) plot visualizes the non-linear relationship between age and survival:

- **X-axis**: Age (continuous predictor)
- **Y-axis**: Hazard Ratio (HR) with reference at the median age
- **Solid line**: Estimated HR across age values
- **Shaded area**: 95% confidence interval
- **Histogram (top)**: Distribution of age in the dataset

Key features:
- The spline with 4 knots (`knot = 4`) allows flexible modeling of non-linear relationships
- Reference point (`ref = "x_median"`) is set at the median age (HR = 1.0)
- The plot shows how the hazard changes relative to the median age
- The relatively flat curve suggests a roughly linear relationship between age and hazard in this dataset

### Predictor Effect Plot

A unified interface for both continuous and categorical predictors:

```{r predictor-effect}
effect_linear <- predictor_effect_plot(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  covars = c("sex"),
  method = "linear",
  add_hist = TRUE,
  save_plot = FALSE
)

effect_linear
```

The linear predictor effect plot shows:

- **X-axis**: Age values
- **Y-axis**: Predicted effect (log hazard ratio)
- **Solid line**: Linear fit with 95% confidence interval (shaded area)
- **Rug plot (bottom)**: Individual data point distribution

This plot assumes a linear relationship between the predictor and outcome, useful for confirming linearity assumptions or presenting simple linear effects.

```{r predictor-effect-categorical}
effect_cat <- predictor_effect_plot(
  data = cancer_clean,
  x = "sex",
  y = "status",
  time = "time",
  covars = c("age", "ph.karno"),
  method = "categorical",
  save_plot = FALSE
)

effect_cat
```

The categorical predictor effect plot displays:

- **X-axis**: Categories of the predictor (Male vs Female)
- **Y-axis**: Hazard Ratio with 95% confidence intervals
- **Reference line**: HR = 1.0 (dashed line)
- **Error bars**: 95% confidence intervals for each category

In this example, females show a significantly lower hazard compared to males (HR ≈ 0.62, 95% CI does not include 1.0), indicating better survival outcomes for female patients after adjusting for age and performance status.

### Customized RCS Plot

```{r custom-effect}
custom_rcs <- rcs_plot(
  data = cancer_clean,
  x = "wt.loss",
  y = "status",
  time = "time",
  covars = c("age", "sex"),
  knot = 5,
  ref = 0,
  add_hist = FALSE,
  trans = "log10",
  y_lim = c(0.5, 5),
  save_plot = FALSE
)

custom_rcs
```

The customized RCS plot demonstrates advanced options:

- **`knot = 5`**: Uses 5 knots for more flexible curve fitting
- **`ref = 0`**: Sets reference point at weight loss = 0 (no weight loss)
- **`trans = "log10"`**: Applies log10 transformation to the y-axis
- **`y_lim = c(0.5, 5)`**: Custom y-axis limits for better visualization
- **`add_hist = FALSE`**: Excludes the histogram for cleaner presentation

This plot examines the relationship between weight loss and survival, with the reference at zero weight loss. The log-scale y-axis helps visualize effect sizes across a wide range.

## Interaction Analysis

### Scanning for Interactions

```{r interaction-scan}
interaction_scan <- interaction_scan(
  data = cancer_clean,
  y = "status",
  time = "time",
  predictors = c("age", "ph.karno", "pat.karno"),
  group_vars = c("sex"),
  save_table = FALSE
)

knitr::kable(interaction_scan, caption = "Interaction Scan Results")
```

### Visualizing Interactions

```{r interaction-plot}
interaction_age_sex <- interaction_plot(
  data = cancer_clean,
  predictor = "age",
  y = "status",
  time = "time",
  group_var = "sex",
  covars = c("ph.karno"),
  save_plot = FALSE
)

interaction_age_sex
```

The interaction plot displays two panels:

**Linear Interaction Plot (`$lin`)**:
- Shows the effect of age on survival separately for males and females
- Different colored lines represent different sex groups
- Parallel lines suggest no interaction; diverging/converging lines suggest interaction
- Shaded areas represent 95% confidence intervals

**RCS Interaction Plot (`$rcs`)**:
- Shows non-linear effects of age within each sex group
- Allows detection of non-linear interactions
- Each curve represents the HR relative to the reference point for that sex

In this example, the lines appear roughly parallel, suggesting no significant interaction between age and sex (confirmed by p-value > 0.05).

### Interaction P-value

```{r interaction-pvalue}
interaction_p <- interaction_p_value(
  data = cancer_clean,
  predictor = "age",
  y = "status",
  time = "time",
  group_var = "sex",
  covars = c("ph.karno")
)

knitr::kable(data.frame(
  Interaction = "age * sex",
  P_Value = round(interaction_p, 4),
  Interpretation = ifelse(interaction_p < 0.05,
                         "Significant interaction",
                         "No significant interaction")
), caption = "Interaction Test Result")
```

## Subgroup Analysis

```{r subgroup-forest}
subgroup_result <- subgroup_forest(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  subgroup_vars = c("sex"),
  covars = c("ph.karno"),
  save_plot = FALSE
)

subgroup_result
```

The subgroup forest plot displays the effect of age on survival stratified by sex:

- **Overall**: Effect estimate for the entire cohort
- **Male**: Effect estimate for male patients only
- **Female**: Effect estimate for female patients only

Each row shows:
- Sample size for that subgroup
- Hazard ratio with 95% confidence interval
- P-value for statistical significance

This visualization helps assess:
- Whether the effect is consistent across subgroups
- Potential effect modification by the stratifying variable
- Heterogeneity between subgroups

In this example, the effect of age appears similar between males and females, with overlapping confidence intervals, suggesting no substantial subgroup differences.

## Model Fitting

For custom analyses, fit models directly with full control:

```{r regression-fit}
fit_result <- regression_fit(
  data = cancer_clean,
  y = "status",
  predictor = "age",
  time = "time",
  covars = c("sex", "ph.karno"),
  returned = "full"
)

knitr::kable(fit_result, caption = "Detailed Model Results")
```

### RCS Model Fitting

```{r rcs-fit}
rcs_fit <- regression_fit(
  data = cancer_clean,
  y = "status",
  predictor = "age",
  time = "time",
  covars = c("sex"),
  rcs_knots = 4,
  returned = "predictor_combined"
)

knitr::kable(rcs_fit, caption = "RCS Model Results")
```

## Complete Analysis Workflow

```{r complete-workflow}
# Step 1: Data preparation
data(cancer, package = "survival")
cancer$dead <- cancer$status == 2

analysis_data <- cancer %>%
  mutate(
    sex = factor(sex, labels = c("Male", "Female")),
    ph.ecog_cat = factor(ph.ecog, levels = 0:3,
                         labels = c("0", "1", "2", "3"))
  ) %>%
  na.omit()

cat("Analysis dataset:", nrow(analysis_data), "patients\n")

# Step 2: Regression scan
scan_results <- regression_scan(
  analysis_data,
  y = "status",
  time = "time",
  predictors = c("age", "sex", "ph.karno", "pat.karno", "wt.loss", "meal.cal"),
  save_table = FALSE
)

top_predictors <- scan_results %>%
  filter(original.pval < 0.05) %>%
  pull(predictor)

cat("Significant predictors:", paste(top_predictors, collapse = ", "), "\n")

knitr::kable(
  scan_results %>%
    filter(predictor %in% top_predictors) %>%
    select(predictor, best.var.trans, original.pval),
  caption = "Key Significant Predictors",
  digits = 4
)

# Step 3: Comprehensive results for key predictor
age_results <- regression_basic_results(
  analysis_data,
  x = "age",
  y = "status",
  time = "time",
  model_covs = list(
    Crude = c(),
    Adjusted = c("sex", "ph.karno", "pat.karno")
  ),
  save_output = FALSE
)

knitr::kable(age_results$table, caption = "Age Effect - Multiple Model Specifications")

# Step 4: Forest plot
numeric_predictors <- top_predictors[top_predictors != "sex"]
forest_plot <- regression_forest(
  analysis_data,
  model_vars = numeric_predictors,
  y = "status",
  time = "time",
  as_univariate = TRUE,
  save_plot = FALSE
)

# Step 5: RCS plot
rcs_result <- rcs_plot(
  analysis_data,
  x = "age",
  y = "status",
  time = "time",
  covars = c("sex", "ph.karno"),
  save_plot = FALSE
)

# Step 6: Subgroup analysis
subgroup_plot <- subgroup_forest(
  analysis_data,
  x = "age",
  y = "status",
  time = "time",
  subgroup_vars = c("sex"),
  covars = c("ph.karno"),
  save_plot = FALSE
)

cat("\n=== Analysis Complete ===\n")
cat(length(top_predictors), "significant predictors identified\n")
```

## Summary

| Function | Purpose |
|----------|---------|
| `regression_scan()` | Automated screening of predictors with multiple transformations |
| `regression_basic_results()` | Comprehensive results with various model specifications |
| `regression_forest()` | Publication-ready forest plots |
| `rcs_plot()` / `predictor_effect_plot()` | Flexible visualization of predictor effects |
| `interaction_scan()` / `interaction_plot()` | Interaction analysis and visualization |
| `subgroup_forest()` | Subgroup analysis with forest plots |
| `regression_fit()` | Direct model fitting with full control |

## Further Reading

- `vignette("data-cleaning")`: Data cleaning and preparation
- `vignette("baseline-table")`: Creating baseline characteristics tables
- `vignette("model-evaluation")`: Model comparison and evaluation
