---
title: "Regression Analysis and Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Analysis and Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

```{r setup}
library(clinpubr)
library(dplyr)
library(survival)
library(ggplot2)
```

## Introduction

The `clinpubr` package provides a comprehensive suite for regression analysis in clinical research, supporting Cox proportional hazards, logistic, and linear regression with publication-ready visualizations. This vignette covers the typical analysis workflow:

1. **Regression Scan** --- Screen predictors with multiple transformations
2. **Comprehensive Results** --- Generate multi-model specification tables
3. **Forest Plots** --- Visualize effect sizes
4. **Predictor Effects** --- RCS plots and effect plots
5. **Interaction Analysis** --- Test and visualize effect modification
6. **Subgroup Analysis** --- Subgroup forest plots

## Data Preparation

We'll use the NCCTG Lung Cancer dataset (228 patients with advanced lung cancer):

```{r load-data}
data(cancer, package = "survival")
cancer$dead <- cancer$status == 2

cancer_clean <- cancer %>%
  mutate(
    sex = factor(sex, labels = c("Male", "Female")),
    ph.ecog_cat = factor(ph.ecog, levels = 0:3,
                         labels = c("0 - Asymptomatic",
                                   "1 - Symptomatic but ambulatory",
                                   "2 - In bed <50% of time",
                                   "3 - In bed >50% of time"))
  ) %>%
  na.omit()

knitr::kable(data.frame(
  Rows = nrow(cancer_clean),
  Columns = ncol(cancer_clean),
  Description = "Complete cases for analysis"
), caption = "Analysis Dataset Summary")

knitr::kable(head(cancer_clean[, c("age", "sex", "ph.ecog", "ph.karno", "time", "status")]),
             caption = "Sample of Analysis Variables")
```

## Regression Scan

`regression_scan()` automatically tests multiple variable transformations for each predictor:

- **Original** (linear) form
- **Logarithmic** transformation (if all values > 0)
- **Categorical** form (median split)
- **RCS** (Restricted Cubic Spline) for non-linear relationships

### Cox Proportional Hazards Model

```{r cox-scan}
cox_scan <- regression_scan(
  data = cancer_clean,
  y = "status",
  time = "time",
  predictors = c("age", "sex", "ph.karno", "pat.karno", "wt.loss", "meal.cal"),
  covars = NULL,
  save_table = FALSE
)

knitr::kable(cox_scan, caption = "Cox Regression Scan Results")
```

The scan reports `original.pval` (linear), `logarithm.pval` (log-transformed), `categorized.pval` (median split), `rcs.overall.pval` (RCS overall), and `best.var.trans` (best transformation).

View significant predictors:

```{r interpret-scan}
significant_predictors <- cox_scan %>%
  filter(original.pval < 0.05 | rcs.overall.pval < 0.05) %>%
  select(predictor, nvalid, best.var.trans, original.pval, rcs.overall.pval)

knitr::kable(significant_predictors,
             caption = "Significant Predictors (p < 0.05)",
             digits = 4)
```

### Logistic Regression Scan

```{r logistic-scan}
logistic_scan <- regression_scan(
  data = cancer_clean,
  y = "dead",
  predictors = c("age", "sex", "ph.karno", "pat.karno", "wt.loss"),
  covars = NULL,
  save_table = FALSE
)

knitr::kable(logistic_scan, caption = "Logistic Regression Scan Results")
```

## Comprehensive Regression Results

In clinical publications, it's standard to present multiple model specifications (Crude, Partially adjusted, Fully adjusted) to show how confounding affects the estimated association.

### Cox Model with Multiple Specifications

```{r cox-basic-results}
cox_results <- regression_basic_results(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  model_covs = list(
    Crude = c(),
    Model1 = c("sex"),
    Model2 = c("sex", "ph.karno", "pat.karno")
  ),
  pers = c(1, 5, 10),
  save_output = FALSE
)

knitr::kable(cox_results$table, caption = "Cox Regression Results for Age")
```

### Logistic Model with Clinical Cut-points

```{r logistic-basic-results}
logistic_results <- regression_basic_results(
  data = cancer_clean,
  x = "ph.karno",
  y = "dead",
  model_covs = list(
    Crude = c(),
    Adjusted = c("age", "sex")
  ),
  factor_breaks = c(80),
  factor_labels = c("<80", ">=80"),
  save_output = FALSE
)

knitr::kable(logistic_results$table, caption = "Logistic Regression Results for Performance Status")
```

### Quantile-Based Categories

When clinical cut-points aren't established, use tertiles for balanced group comparison:

```{r custom-transformations}
custom_results <- regression_basic_results(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  quantile_breaks = c(0.33, 0.67),
  quantile_labels = c("Youngest Third", "Middle Third", "Oldest Third"),
  label_with_range = TRUE,
  save_output = FALSE
)

knitr::kable(custom_results$table, caption = "Cox Regression Results - Age Tertiles")
```

## Forest Plots

### Univariate Forest Plot

```{r univariate-forest}
uni_forest <- regression_forest(
  data = cancer_clean,
  model_vars = c("age", "ph.karno", "pat.karno", "wt.loss"),
  y = "status",
  time = "time",
  as_univariate = TRUE,
  save_plot = FALSE
)

uni_forest
```

### Multivariate Forest Plot

```{r multivariate-forest}
multi_forest <- regression_forest(
  data = cancer_clean,
  model_vars = list(
    Model1 = c("age"),
    Model2 = c("age", "ph.karno"),
    Model3 = c("age", "ph.karno", "pat.karno")
  ),
  y = "status",
  time = "time",
  as_univariate = FALSE,
  save_plot = FALSE
)

multi_forest
```

### Customizing Forest Plots

```{r custom-forest}
custom_forest <- regression_forest(
  data = cancer_clean,
  model_vars = c("age", "ph.karno", "pat.karno", "wt.loss", "meal.cal"),
  y = "status",
  time = "time",
  as_univariate = TRUE,
  show_vars = c("age", "ph.karno", "pat.karno"),
  est_nsmall = 2,
  p_nsmall = 4,
  save_plot = FALSE
)

custom_forest
```

## Predictor Effect Visualization

### RCS Plots

RCS plots visualize non-linear relationships while preserving the continuous nature of the data:

```{r rcs-plot}
rcs_age <- rcs_plot(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  covars = c("sex", "ph.karno"),
  knot = 4,
  ref = "x_median",
  add_hist = TRUE,
  save_plot = FALSE
)

rcs_age
```

### Predictor Effect Plot

A unified interface for both continuous and categorical predictors:

```{r predictor-effect}
effect_linear <- predictor_effect_plot(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  covars = c("sex"),
  method = "linear",
  add_hist = TRUE,
  save_plot = FALSE
)

effect_linear
```

```{r predictor-effect-categorical}
effect_cat <- predictor_effect_plot(
  data = cancer_clean,
  x = "sex",
  y = "status",
  time = "time",
  covars = c("age", "ph.karno"),
  method = "categorical",
  save_plot = FALSE
)

effect_cat
```

### Customized RCS Plot

```{r custom-effect}
custom_rcs <- rcs_plot(
  data = cancer_clean,
  x = "wt.loss",
  y = "status",
  time = "time",
  covars = c("age", "sex"),
  knot = 5,
  ref = 0,
  add_hist = FALSE,
  trans = "log10",
  y_lim = c(0.5, 5),
  save_plot = FALSE
)

custom_rcs
```

## Interaction Analysis

### Scanning for Interactions

```{r interaction-scan}
interaction_scan <- interaction_scan(
  data = cancer_clean,
  y = "status",
  time = "time",
  predictors = c("age", "ph.karno", "pat.karno"),
  group_vars = c("sex"),
  save_table = FALSE
)

knitr::kable(interaction_scan, caption = "Interaction Scan Results")
```

### Visualizing Interactions

```{r interaction-plot}
interaction_age_sex <- interaction_plot(
  data = cancer_clean,
  predictor = "age",
  y = "status",
  time = "time",
  group_var = "sex",
  covars = c("ph.karno"),
  save_plot = FALSE
)

interaction_age_sex
```

### Interaction P-value

```{r interaction-pvalue}
interaction_p <- interaction_p_value(
  data = cancer_clean,
  predictor = "age",
  y = "status",
  time = "time",
  group_var = "sex",
  covars = c("ph.karno")
)

knitr::kable(data.frame(
  Interaction = "age * sex",
  P_Value = round(interaction_p, 4),
  Interpretation = ifelse(interaction_p < 0.05,
                         "Significant interaction",
                         "No significant interaction")
), caption = "Interaction Test Result")
```

## Subgroup Analysis

```{r subgroup-forest}
subgroup_result <- subgroup_forest(
  data = cancer_clean,
  x = "age",
  y = "status",
  time = "time",
  subgroup_vars = c("sex"),
  covars = c("ph.karno"),
  save_plot = FALSE
)

subgroup_result
```

## Model Fitting

For custom analyses, fit models directly with full control:

```{r regression-fit}
fit_result <- regression_fit(
  data = cancer_clean,
  y = "status",
  predictor = "age",
  time = "time",
  covars = c("sex", "ph.karno"),
  returned = "full"
)

knitr::kable(fit_result, caption = "Detailed Model Results")
```

### RCS Model Fitting

```{r rcs-fit}
rcs_fit <- regression_fit(
  data = cancer_clean,
  y = "status",
  predictor = "age",
  time = "time",
  covars = c("sex"),
  rcs_knots = 4,
  returned = "predictor_combined"
)

knitr::kable(rcs_fit, caption = "RCS Model Results")
```

## Complete Analysis Workflow

```{r complete-workflow}
# Step 1: Data preparation
data(cancer, package = "survival")
cancer$dead <- cancer$status == 2

analysis_data <- cancer %>%
  mutate(
    sex = factor(sex, labels = c("Male", "Female")),
    ph.ecog_cat = factor(ph.ecog, levels = 0:3,
                         labels = c("0", "1", "2", "3"))
  ) %>%
  na.omit()

cat("Analysis dataset:", nrow(analysis_data), "patients\n")

# Step 2: Regression scan
scan_results <- regression_scan(
  analysis_data,
  y = "status",
  time = "time",
  predictors = c("age", "sex", "ph.karno", "pat.karno", "wt.loss", "meal.cal"),
  save_table = FALSE
)

top_predictors <- scan_results %>%
  filter(original.pval < 0.05) %>%
  pull(predictor)

cat("Significant predictors:", paste(top_predictors, collapse = ", "), "\n")

knitr::kable(
  scan_results %>%
    filter(predictor %in% top_predictors) %>%
    select(predictor, best.var.trans, original.pval),
  caption = "Key Significant Predictors",
  digits = 4
)

# Step 3: Comprehensive results for key predictor
age_results <- regression_basic_results(
  analysis_data,
  x = "age",
  y = "status",
  time = "time",
  model_covs = list(
    Crude = c(),
    Adjusted = c("sex", "ph.karno", "pat.karno")
  ),
  save_output = FALSE
)

knitr::kable(age_results$table, caption = "Age Effect - Multiple Model Specifications")

# Step 4: Forest plot
numeric_predictors <- top_predictors[top_predictors != "sex"]
forest_plot <- regression_forest(
  analysis_data,
  model_vars = numeric_predictors,
  y = "status",
  time = "time",
  as_univariate = TRUE,
  save_plot = FALSE
)

# Step 5: RCS plot
rcs_result <- rcs_plot(
  analysis_data,
  x = "age",
  y = "status",
  time = "time",
  covars = c("sex", "ph.karno"),
  save_plot = FALSE
)

# Step 6: Subgroup analysis
subgroup_plot <- subgroup_forest(
  analysis_data,
  x = "age",
  y = "status",
  time = "time",
  subgroup_vars = c("sex"),
  covars = c("ph.karno"),
  save_plot = FALSE
)

cat("\n=== Analysis Complete ===\n")
cat(length(top_predictors), "significant predictors identified\n")
```

## Summary

| Function | Purpose |
|----------|---------|
| `regression_scan()` | Automated screening of predictors with multiple transformations |
| `regression_basic_results()` | Comprehensive results with various model specifications |
| `regression_forest()` | Publication-ready forest plots |
| `rcs_plot()` / `predictor_effect_plot()` | Flexible visualization of predictor effects |
| `interaction_scan()` / `interaction_plot()` | Interaction analysis and visualization |
| `subgroup_forest()` | Subgroup analysis with forest plots |
| `regression_fit()` | Direct model fitting with full control |

## Further Reading

- `vignette("data-cleaning")`: Data cleaning and preparation
- `vignette("baseline-table")`: Creating baseline characteristics tables
- `vignette("model-evaluation")`: Model comparison and evaluation
